---
title: "NACCADRC Basic Summaries"
format: 
   html:
     toc: true
     toc_depth: 3
     code_fold: true
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{NACCADRC Basic Summaries}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---
  
```{r packages, message = FALSE}
#| code-fold: true

library(NACCADRC)
library(ggplot2)
library(tidyverse)
library(nlme)
library(data.table)
library(gtsummary)
library(UpSetR)
library(patchwork)
library(sva)

data(mrisbm, taupetnpdka, phc_cognition, amyloidpetgaain, 
uds_ftldlbd, clariti_edc, uds_mri)

theme_set(theme_minimal())

scale_colour_discrete <-
  function(...) ggsci::scale_color_jama()
scale_fill_discrete <-
  function(...) ggsci::scale_fill_jama()

cohort_color <- c(CLARiTI = "#C02126", SCAN = "#173058", "Mixed protocol" = "#42B540FF", "SCAN MP" = "#42B540FF")
```

```{r settings, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 4, 
  out.width = "100%",
  class.source = "fold-hide",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)

options(digits = 2, width = 150)

# Package Versions
pkgName <- "NACCADRC"
pkgFilename <- paste0('NACCADRC_', packageVersion('NACCADRC'), '.tar.gz')
```

# Merge key imaging and PHC cognitive data

```{r merge-data}
#| code-fold: true

## Combine hippocampal volume from SCAN and Mixed Protocol (UDS) ----
mri_hipp <- mrisbm %>%          # SCAN volumes
  select(SOURCE, NACCID, DATE = SCANDT, HIPPOCAMPUS, ICV = CEREBRUMTCV) %>%
  bind_rows(uds_mri %>%         # Mixed Protocol (UDS) volumes
    select(NACCID, MRIYR, MRIMO, MRIDY, HIPPOCAMPUS = HIPPOVOL, ICV = NACCICV) %>%
    filter(HIPPOCAMPUS != 88.8888, ICV != 9999.999, !is.na(HIPPOCAMPUS)) %>%
    mutate(
      DATE = as.IDate(paste(MRIYR, MRIMO, MRIDY, sep='-')),
      SOURCE = 'Mixed protocol')) %>%
  distinct(NACCID, DATE, .keep_all = TRUE)

NACCETPR_levels <- names(rev(sort(table(uds_ftldlbd$NACCETPR))))

## Combine hipp volume, tau PET, amyloid PET, cognition, and demographics ----
dd_atn <- clariti_edc %>% 
  select(NACCID) %>% distinct() %>% mutate(`CLARiTI EDC` = 'Yes') %>%
  full_join(mri_hipp %>%                 # Hippocampal volumes
      select(NACCID, MRI_SOURCE = SOURCE, DATE, HIPPOCAMPUS, ICV),
    by = 'NACCID') %>%
  full_join(taupetnpdka %>%
      arrange(NACCID, SCANDATE, desc(PROCESSDATE)) %>%
      distinct(NACCID, SCANDATE, .keep_all = TRUE) %>%
      select(NACCID, DATE = SCANDATE, Tau_PET = META_TEMPORAL_SUVR, 
        Tau_TRACER = TRACER, Tau_SOURCE = SOURCE) %>%
      distinct(NACCID, DATE, .keep_all = TRUE),
    by = c('NACCID', 'DATE')) %>%
  full_join(amyloidpetgaain %>%          # Amyloid PET
      arrange(NACCID, SCANDATE, desc(PROCESSDATE)) %>%
      distinct(NACCID, SCANDATE, .keep_all = TRUE) %>%
      select(NACCID, DATE = SCANDATE, AMYLOID_STATUS, CENTILOIDS, 
        Amyloid_TRACER = TRACER, Amyloid_SOURCE = SOURCE) %>%
      distinct(NACCID, DATE, .keep_all = TRUE),
    by = c('NACCID', 'DATE')) %>%
  full_join(phc_cognition %>%            # Harmonized cognition
      select(NACCID, NACCVNUM, PHC_MEM, PHC_EXF, PHC_LAN) %>%
      distinct(NACCID, NACCVNUM, .keep_all = TRUE) %>%
      left_join(uds_ftldlbd %>%          # UDS visit dates
          select(NACCID, NACCVNUM, DATE = VISITDATE),
        by = c('NACCID', 'NACCVNUM')) %>%
      distinct(NACCID, DATE, .keep_all = TRUE),
    by = c('NACCID', 'DATE')) 

stopifnot(with(dd_atn, !any(duplicated(paste(NACCID, DATE)))))

dd <- dd_atn %>%
  full_join(uds_ftldlbd %>%              # UDS diagnosis, etiology over time
      filter(NACCID %in% c(dd_atn$NACCID, clariti_edc$NACCID)) %>%
      mutate(
        NACCETPR = factor(NACCETPR, levels = NACCETPR_levels)) %>%
      select(NACCID, DATE = VISITDATE, NACCUDSD, NACCETPR, PARK) %>%
      distinct(NACCID, DATE, .keep_all = TRUE),
    by = c('NACCID', 'DATE')) %>%
  left_join(uds_ftldlbd %>%              # UDS demographics
      mutate(BIRTHDATE = as.IDate(paste(BIRTHYR, BIRTHMO, 15, sep = '-'))) %>%
      filter(!is.na(BIRTHDATE)) %>%
      arrange(NACCID, NACCVNUM) %>%
      select(NACCID, BIRTHDATE, RACE, SEX = NACCSEX, EDUC, HISPANIC = NACCHISP) %>%
      group_by(NACCID) %>%               # Carry information back/forward
      tidyr::fill(.direction = "updown") %>% # to impute missing data
      ungroup() %>%
      filter(!duplicated(NACCID)),       # One row of demographics per NACCID
    by = 'NACCID') %>%
  arrange(NACCID, DATE) %>%
  group_by(NACCID) %>%                   # Carry Dx information forward/back
  tidyr::fill(all_of(c("NACCUDSD", "NACCETPR", "PARK")), .direction = "downup") %>%
  mutate(ICV = mean(ICV, na.rm = TRUE)) %>% # Average ICV over visits
  ungroup() %>%
  mutate(
    Age = as.numeric(DATE - BIRTHDATE)/365.25,
    Etiology = case_when(                     # Simplified/collapse diagnosis
      NACCETPR == "Alzheimer's disease (AD)" ~ 'AD',
      NACCETPR == 'Lewy body disease (LBD)' | PARK == 'Yes' ~ 'LBD',
      NACCETPR %in% c("FTLD, other", "FTLD with motor neuron disease (e.g., ALS)") ~ 
        "FTLD (any)",
      NACCETPR == "Vascular brain injury or vascular dementia including stroke" ~ "Vascular",
      NACCETPR == 'Not applicable, not cognitively impaired' ~ 'Not impaired',
      TRUE ~ 'Other') %>%
      factor(levels = c('Not impaired', 'LBD', "FTLD (any)", 'AD', "Vascular", 'Other')),
    NACCUDSD = case_when(
      is.na(NACCUDSD) ~ 'Unknown',
      TRUE ~ NACCUDSD) %>%
      factor(levels = c("Normal cognition", "Impaired-not-MCI", 
        "MCI", "Dementia", 'Unknown')))

stopifnot(with(dd, !any(duplicated(paste(NACCID, DATE)))))

CLARiTI_id <- dd %>%
  filter(MRI_SOURCE == 'CLARiTI' | Tau_SOURCE == 'CLARiTI' | Amyloid_SOURCE == 'CLARiTI') %>%
  pull(NACCID) %>% unique()

SCAN_id <- dd %>%
  filter(MRI_SOURCE == 'SCAN' | Tau_SOURCE == 'SCAN' | Amyloid_SOURCE == 'SCAN') %>%
  pull(NACCID) %>% unique() %>%
  setdiff(CLARiTI_id)

Mixed_id <- dd %>%
  filter(MRI_SOURCE == 'Mixed protocol' | Tau_SOURCE == 'SCAN MP' | 
    Amyloid_SOURCE == 'SCAN MP') %>%
  pull(NACCID) %>% unique() %>%
  setdiff(c(CLARiTI_id, SCAN_id))

# harmonize tau PET data ----
tmp <- dd %>% filter(!is.na(Tau_PET) & !is.na(Age)) %>%
  mutate(across(where(is.factor), fct_drop))
set.seed(42)
tmp$Tau_PET_ComBat <- sva::ComBat(
  dat=rbind(tmp$Tau_PET, jitter(tmp$Tau_PET)), # ComBat needs two rows 
  batch=tmp$Tau_TRACER, 
  mod=model.matrix(~Age + SEX, tmp), par.prior=TRUE, prior.plots=FALSE)[1,]

dd <- dd %>%
  left_join(tmp %>% select(NACCID, DATE, Tau_PET_ComBat), 
    by = c('NACCID', 'DATE'))

# x-sectional data ----
dd_cross <- dd %>%
  group_by(NACCID) %>%
  fill(everything(), .direction = "updown") %>%
  filter(!duplicated(NACCID)) %>%
  ungroup() %>%
  mutate(    # add labels for tables
    Cohort = case_when(
      NACCID %in% clariti_edc$NACCID ~ 'CLARiTI',
      NACCID %in% CLARiTI_id ~ 'CLARiTI',
      NACCID %in% SCAN_id ~ 'SCAN',
      NACCID %in% Mixed_id ~ 'Mixed protocol',
      TRUE ~ 'UDS only') %>%
      factor(levels = c('CLARiTI', 'SCAN', 'Mixed protocol', 'UDS only')),
    PHC_MEM = structure(PHC_MEM, label = 'Harmonized Memory'),
    PHC_EXF = structure(PHC_EXF, label = 'Harmonized Exec Function'), 
    PHC_LAN = structure(PHC_LAN, label = 'Harmonized Language'),
    EDUC = structure(EDUC, label = 'Education (yrs)'),
    Age = structure(Age, label = 'Age (yrs)'),
    SEX = structure(SEX, label = 'Sex'),
    RACE = structure(RACE, label = 'Race'),
    HISPANIC = structure(HISPANIC, label = 'Hispanic'),
    AMYLOID_STATUS = structure(AMYLOID_STATUS, label = 'Amyloid status'),
    CENTILOIDS = structure(CENTILOIDS, label = 'Amyloid (CL)'),
    HIPPOCAMPUS = structure(HIPPOCAMPUS, label = 'Hippocampal volume (mm3)'),
    Tau_PET = structure(Tau_PET, label = 'Tau PET (SUVR)'),
    Tau_PET_ComBat = structure(Tau_PET_ComBat, label = 'Tau PET (SUVR)'),
    Tau_TRACER = structure(Tau_TRACER, label = 'Tau PET tracer'),
    NACCETPR = structure(NACCETPR, label = 'Primary etiologic diagnosis'),
    NACCUDSD = structure(NACCUDSD, label = 'Cognitive status at UDS visit'))
```

# Summarize data collection

## CLARiTI imaging distributions

```{r fig-clariti-imaging}
#| code-fold: true
#| fig-cap: "Distribution of imaging summaries among CLARiTI participants with known UDS cognitive status, amyloid PET (CL), hippocampal volume (mm$^3$), and tau PET (SUVR). Tau PET is shown as SUVR harmonized across tracers by ComBat."

dd_cross %>%
  filter(Cohort == 'CLARiTI') %>%
  filter(NACCUDSD != 'Unknown') %>%
  mutate(
    Diagnosis = case_when(
      NACCUDSD == "Normal cognition" ~ "NC",
      NACCETPR == "Alzheimer's disease (AD)" ~ "CI-AD",
      TRUE ~ "CI-nonAD") %>% factor(c("NC", "CI-AD", "CI-nonAD"))) %>%
  filter(!is.na(Diagnosis), !is.na(CENTILOIDS), !is.na(Tau_PET_ComBat),
    !is.na(HIPPOCAMPUS)) %>%
  select(NACCID, Diagnosis, 
    `Amyloid (CL)` = CENTILOIDS, 
    `Tau PET MTL (SUVR)` = Tau_PET_ComBat, 
    `Hippocampal volume (mm3)` = HIPPOCAMPUS) %>%
  pivot_longer(`Amyloid (CL)`:`Hippocampal volume (mm3)`, 
    names_to = 'Measure', values_to = 'Value') %>%
ggplot(aes(x = Diagnosis, y = Value)) +
  geom_violin(fill = "gray80", color = "gray50") +
  ggbeeswarm::geom_beeswarm(color = "blue", alpha = 0.5) +
  facet_wrap(vars(Measure), nrow = 1, scales = 'free', strip.position = "left") +
  labs(x = "", y = "") +
  theme(legend.position = "none", strip.placement = "outside")
```

## Bar charts

```{r fig-clariti-barplots}
#| code-fold: true
#| fig-cap: "Number of CLARiTI pariticpants by etiology and UDS clinical diagnosis."

plot_data <- dd_cross %>% 
  filter(Cohort == 'CLARiTI') %>%
  filter(NACCUDSD != 'Unknown') %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology) %>%
  summarise(n = n(), .groups = "drop")
ggplot(plot_data, aes(x = n, y = Etiology)) +
  geom_bar(stat = "identity", fill = cohort_color["CLARiTI"]) +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  expand_limits(x = max(plot_data$n) * 1.2) +
  facet_wrap(vars(NACCUDSD)) +
  labs(y = "", x = "Count (N)")
```

```{r fig-all-imaging-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with any imaging data including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

count_data <- dd_cross %>% 
  filter(NACCUDSD != 'Unknown', Cohort != 'UDS only') %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology) %>%
  summarise(n = n(), .groups = "drop")

dd_cross %>% 
  filter(NACCUDSD != 'Unknown', Cohort != 'UDS only') %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology, Cohort) %>%
  summarise(n = n(), .groups = "drop") %>%
ggplot(aes(x = n, y = Etiology)) +
  geom_bar(aes(fill = Cohort), stat = "identity") +
  geom_text(data = count_data, aes(label = n), 
    hjust = -0.2, size = 3) +
  expand_limits(x = max(count_data$n) * 1.2) +
  facet_wrap(vars(NACCUDSD)) +
  labs(y = "", x = "Count (N)") +
  scale_fill_manual(values = cohort_color)
```

```{r fig-amy-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with amyloid PET, including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

count_data <- dd_cross %>% 
  filter(NACCUDSD != 'Unknown' & !is.na(CENTILOIDS)) %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology) %>%
  summarise(n = n(), .groups = "drop")

dd_cross %>% 
  filter(NACCUDSD != 'Unknown' & !is.na(CENTILOIDS)) %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology, Cohort) %>%
  summarise(n = n(), .groups = "drop") %>%
ggplot(aes(x = n, y = Etiology)) +
  geom_bar(aes(fill = Cohort), stat = "identity") +
  geom_text(data = count_data, aes(label = n), 
    hjust = -0.2, size = 3) +
  expand_limits(x = max(count_data$n) * 1.2) +
  facet_wrap(vars(NACCUDSD)) +
  labs(y = "", x = "Count (N)") +
  scale_fill_manual(values = cohort_color)
```

```{r fig-tau-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with tau PET, including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

count_data <- dd_cross %>% 
  filter(NACCUDSD != 'Unknown' & !is.na(Tau_PET)) %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology) %>%
  summarise(n = n(), .groups = "drop")

dd_cross %>% 
  filter(NACCUDSD != 'Unknown' & !is.na(Tau_PET)) %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  group_by(NACCUDSD, Etiology, Cohort) %>%
  summarise(n = n(), .groups = "drop") %>%
ggplot(aes(x = n, y = Etiology)) +
  geom_bar(aes(fill = Cohort), stat = "identity") +
  geom_text(data = count_data, aes(label = n), 
    hjust = -0.2, size = 3) +
  expand_limits(x = max(count_data$n) * 1.2) +
  facet_wrap(vars(NACCUDSD)) +
  labs(y = "", x = "Count (N)") +
  scale_fill_manual(values = cohort_color)
```

## UpSet plots

```{r fig-clariti-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of CLARiTI participants."

dd_upset <- dd_cross %>%
  mutate(
    `CLARiTI EDC` = case_when(`CLARiTI EDC` == 'Yes' ~ 1, TRUE ~ 0),
    Diagnosis = case_when(!is.na(NACCUDSD) &  NACCUDSD != "Unknown" ~ 1, TRUE ~ 0),
    MRI = case_when(!is.na(HIPPOCAMPUS) ~ 1, TRUE ~ 0),
    `Amyloid PET` = case_when(!is.na(CENTILOIDS) ~ 1, TRUE ~ 0),
    `Tau PET` = case_when(!is.na(Tau_PET) ~ 1, TRUE ~ 0)) %>%
  select(NACCID, Cohort, `CLARiTI EDC`, Diagnosis:`Tau PET`) %>%
  as.data.frame()

# Output list CLARiTI participants with missing data for QC ----
if(FALSE){
  qc_dir <- file.path('..', 'reports', 'qc')
  dir.create(qc_dir, recursive = TRUE)
  dd_upset %>% 
    rowwise() %>%
    filter(`CLARiTI EDC` == 1, sum(c_across(MRI:`Tau PET`)) == 0) %>%
    write.csv(file.path(qc_dir, "clariti_edc_no_imaging_participants.csv"), row.names = FALSE)
  
  dd_upset %>% 
    rowwise() %>%
    filter(`CLARiTI EDC` == 1, sum(c_across(Diagnosis:`Tau PET`)) == 0) %>%
    write.csv(file.path(qc_dir, "clariti_edc_only_participants.csv"), row.names = FALSE)
  
  dd_upset %>% 
    filter(Cohort == 'CLARiTI') %>%
    write.csv(file.path(qc_dir, "clariti_participants.csv"), row.names = FALSE)
}

# UpSet plot of CLARiTI participants ----
upset(subset(dd_upset, Cohort == 'CLARiTI'), nsets = 7, nintersects = 30, mb.ratio = c(0.5, 0.5),
  order.by = c("freq", "degree"), decreasing = c(TRUE,FALSE))
```

```{r fig-clariti-scan-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of CLARiTI and SCAN participants."

upset(subset(dd_upset, Cohort %in% c('CLARiTI', 'SCAN')), 
  nsets = 7, nintersects = 30, mb.ratio = c(0.5, 0.5),
  order.by = c("freq", "degree"), decreasing = c(TRUE,FALSE))
```

```{r fig-any-imaging-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of participants with any imaging."

upset(subset(dd_upset, Cohort %in% c('CLARiTI', 'SCAN', 'Mixed protocol')), 
  nsets = 7, nintersects = 30, mb.ratio = c(0.5, 0.5),
  order.by = c("freq", "degree"), decreasing = c(TRUE,FALSE))
```

```{r fig-all-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of all participants."

upset(dd_upset, nsets = 7, nintersects = 30, mb.ratio = c(0.5, 0.5),
  order.by = c("freq", "degree"), decreasing = c(TRUE,FALSE))
```

## Scan counts

```{r tbl-scan-counts, results = 'asis'}
#| code-fold: true

tmp <- dd %>% 
  select(NACCID, CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET` = CENTILOIDS, `Volumetric MRI` = HIPPOCAMPUS, 
    `Tau PET` = Tau_PET_ComBat) %>%
  group_by(NACCID) %>%
  pivot_longer(`Amyloid PET`:`Tau PET`) %>%
  filter(!is.na(value)) %>%
  mutate(Type = factor(name, levels = c(
    "Amyloid PET", "Tau PET", "Volumetric MRI"))) %>%
  group_by(Type, NACCID) %>%
  summarise(`Serial scans` = n())

with(tmp, table(Type, `Serial scans`)) %>% 
  knitr::kable(caption = "Number of individuals who have received the given number serial scans for each scan type.")
```

## Cummulative scans by time from first scan

```{r fig-cummulative-scans}
#| code-fold: true
#| fig-cap: "Cumulative scans by time from first scan of each scan type."

dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET` = CENTILOIDS, `Volumetric MRI` = HIPPOCAMPUS, 
    `Tau PET` = Tau_PET_ComBat) %>%
  pivot_longer(`Amyloid PET`:`Tau PET`) %>%
  filter(!is.na(value), !is.na(Age), !is.na(NACCUDSD)) %>%
  mutate(Type = factor(name, levels = c(
    "Amyloid PET", "Tau PET", "Volumetric MRI"))) %>%
  group_by(NACCID, Type) %>%
  mutate(
    Age0 = min(Age, na.rm = TRUE),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down") %>%
  group_by(Type, Years) %>%
  summarise(count = n()) %>%
  mutate(`Cumulative count` = cumsum(count)) %>%
ggplot(aes(x=Years, y=`Cumulative count`, color = Type)) +
  geom_line() +
  xlab('Years from first scan of each type')
```

# Baseline characteristics

## CLARiTI participants

```{r tbl-clariti-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(Cohort == 'CLARiTI') %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("CLARiTI EDC", "Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(
    all_continuous() ~ c(
      "{mean} ({sd})",
      "{median} ({p25}, {p75})",
      "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of CLARiTI participants by baseline UDS diagnosis. Note CLARiTI participants are those with a NACCID in clariti_edc or any of the CLARiTI imaging summary files.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with any imaging

```{r tbl-any-imaging-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>%
    filter(Cohort != 'UDS only') %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of participants with any imaging by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## All participants

```{r tbl-all-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with hippocampal volumes

```{r tbl-mri-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(HIPPOCAMPUS)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"
  ),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", 
    "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants with MRI data by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with tau PET

```{r tbl-tau-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(Tau_PET)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants with tau PET data by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with amyloid PET

```{r tbl-amyloid-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(CENTILOIDS)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"
  ),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of NACC ADRC participants with amyloid PET data by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

# Summary plots

## LOESS Trends

```{r loess, message=FALSE}
#| code-fold: true
#| fig-cap: "LOESS plots. For the bottom plot, memory scores are imputed using linear interpolation (when possible) and a linear mixed effects model (when interpolation was not possible) to allow for plotting against biomarker values. The linear mixed effects model included fixed effects for age (as a cubic polynomial) by initial diagnosis. Random effects included random intercepts and slopes for each participant."

pd1 <- dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET (CL)` = CENTILOIDS, `Hipp. volume` = HIPPOCAMPUS, 
    `Tau PET (MTL SUVR)` = Tau_PET_ComBat) %>%
  group_by(NACCID) %>%
  mutate(
    Age0 = min(Age),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down") %>%
  pivot_longer(`Amyloid PET (CL)`:`Tau PET (MTL SUVR)`) %>%
  filter(!is.na(value), !is.na(Years), !is.na(NACCUDSD)) %>%
  mutate(name = factor(name, levels = c(
    "Amyloid PET (CL)", "Tau PET (MTL SUVR)", "Hipp. volume")))

pd2_0 <- dd %>% 
  select(NACCID, Age, Memory = PHC_MEM, Etiology, NACCUDSD, NACCETPR,
    CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  group_by(NACCID) %>%
  mutate(
    Age0 = min(Age),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down") %>%
  mutate(
    Dx0 = `Initial Dx`,
    Memory_int = zoo::na.approx(Memory, na.rm = FALSE)) %>%
  select(NACCID, Age, `Initial Dx`, Dx0, Memory, Memory_int, Etiology, 
    NACCUDSD, NACCETPR, CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  filter(!is.na(Age), !is.na(Dx0)) %>%
  ungroup()

fit_mem <- lme(Memory ~ I(Age^3)*Dx0, 
  random = ~ Age | NACCID, data = pd2_0, na.action = na.omit)
pd2_0$Memory_lme <- predict(fit_mem, newdata = pd2_0 %>% select(-Memory)) %>%
  as.numeric()

pd2 <- pd2_0 %>%
  mutate(
    Memory_i = case_when(
      !is.na(Memory) ~ Memory,
      !is.na(Memory_int) ~ Memory_int,
      !is.na(Memory_lme) ~ Memory_lme)) %>%
  rename(
    `Amyloid PET (CL)` = CENTILOIDS, `Hipp. volume` = HIPPOCAMPUS, 
    `Tau PET (MTL SUVR)` = Tau_PET_ComBat) %>%
  pivot_longer(`Amyloid PET (CL)`:`Tau PET (MTL SUVR)`) %>%
  filter(!is.na(value), !is.na(Memory_i), !is.na(NACCUDSD)) %>%
  mutate(name = factor(name, levels = c(
    "Amyloid PET (CL)", "Tau PET (MTL SUVR)", "Hipp. volume")))

p1 <- ggplot(pd1, aes(x=Age, y=value, color = `Initial Dx`)) +
  facet_wrap(vars(name), scales = 'free_y', ncol = 3, strip.position = "left") +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", fx = TRUE, k = 1)) +
  coord_cartesian(xlim = c(50, 90)) +
  ylab('') +
  theme(strip.placement = "outside")

p2 <- ggplot(pd2, aes(x=Memory_i, y=value, color = `Initial Dx`)) +
  facet_wrap(vars(name), scales = 'free_y', ncol = 3, strip.position = "left") +
  geom_smooth(method = 'gam', 
    formula = y ~ s(x, bs = "cs", k = 2, sp = 1)) +
  coord_cartesian(xlim = c(-1.5, 1.5)) +
  ylab('Amyloid PET (CL)') +
  xlab('Harmonized Memory (imputed)') +
  scale_x_reverse() +
  ylab('') +
  theme(strip.placement = "outside")

p1 / p2 + plot_layout(guides = "collect")
```

## Spaghetti plots

```{r Spaghetti-centiloids}
#| code-fold: true
#| fig-cap: "Spaghetti plot of amyloid PET."

pd <- dd %>% 
  select(NACCID, SOURCE = Amyloid_SOURCE, Etiology, Age, NACCUDSD, NACCETPR, CENTILOIDS) %>%
  filter(!is.na(CENTILOIDS), !is.na(Age)) %>%
  group_by(NACCID) %>%
  mutate(
    Age0 = min(Age),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down")

ggplot(pd, aes(x = Years, y = CENTILOIDS)) +
  geom_point(aes(color = SOURCE, shape = Etiology), alpha = 0.5) +
  geom_line(aes(group = NACCID), alpha=0.1) +
  geom_density(aes(y = CENTILOIDS, 
    x = after_stat(-scaled)), 
    color = "darkgray", fill = "gray", alpha = 0.3, 
    orientation = "y", inherit.aes = FALSE) +
  scale_x_continuous(labels = function(x) ifelse(x < 0, "", x)) +
  facet_grid(. ~ `Initial Dx`, scales = 'free_x') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  ylab('Amyloid PET (CL)') +
  scale_color_manual(values = cohort_color)
```

```{r Spaghetti-tau-pet}
#| code-fold: true
#| fig-cap: "Spaghetti plot of tau PET."

pd <- dd %>% 
  select(NACCID, SOURCE = Tau_SOURCE, Etiology, Age, NACCUDSD, NACCETPR, Tau_PET_ComBat) %>%
  filter(!is.na(Tau_PET_ComBat), !is.na(Age)) %>%
  group_by(NACCID) %>%
  mutate(
    Age0 = min(Age),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down")

ggplot(pd, aes(x=Years, y=Tau_PET_ComBat)) +
  geom_point(aes(color=SOURCE, shape = Etiology), alpha=1) +
  geom_line(aes(group = NACCID), alpha=0.1) +
  geom_density(aes(y = Tau_PET_ComBat, 
    x = after_stat(-scaled)), 
    color = "darkgray", fill = "gray", alpha = 0.3, 
    orientation = "y", inherit.aes = FALSE) +
  scale_x_continuous(labels = function(x) ifelse(x < 0, "", x)) +
  facet_grid(. ~ `Initial Dx`, scales = 'free_x') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  ylab('Tau PET (MTL SUVR)') +
  scale_color_manual(values = cohort_color)
```

```{r Spaghetti-hippocampus}
#| code-fold: true
#| fig-cap: "Spaghetti plot of hippocampal volumes."

pd <- dd %>% 
  select(NACCID, SOURCE = MRI_SOURCE, Etiology, Age, NACCUDSD, NACCETPR, HIPPOCAMPUS) %>%
  filter(!is.na(HIPPOCAMPUS), !is.na(Age)) %>%
  group_by(NACCID) %>%
  mutate(
    Age0 = min(Age),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down")

ggplot(pd, aes(x=Years, y=HIPPOCAMPUS)) +
  geom_point(aes(color=SOURCE, shape = Etiology), alpha=1) +
  geom_line(aes(group = NACCID), alpha=0.1) +
  geom_density(aes(y = HIPPOCAMPUS, 
    x = after_stat(-scaled)), 
    color = "darkgray", fill = "gray", alpha = 0.3, 
    orientation = "y", inherit.aes = FALSE) +
  scale_x_continuous(labels = function(x) ifelse(x < 0, "", x)) +
  facet_grid(. ~ `Initial Dx`, scales = 'free_x') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  ylab('Hippocampal volume') +
  scale_color_manual(values = cohort_color)
```

```{r Spaghetti-cog}
#| code-fold: true
#| fig-cap: "Spaghetti plot of harmonized cognitive scores."

pd <- dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    PHC_MEM, PHC_EXF, PHC_LAN) %>%
  rename(Memory = PHC_MEM, `Exec. Function` = PHC_EXF, Language = PHC_LAN) %>%
  group_by(NACCID) %>%
  mutate(
    Age0 = min(Age),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down") %>%
  pivot_longer(Memory:Language) %>%
  filter(!is.na(value), !is.na(Years), !is.na(NACCUDSD))

ggplot(pd, aes(x=Years, y=value)) +
  geom_point(aes(color = Etiology), alpha=0.1) +
  geom_density(aes(y = value, 
    x = after_stat(-scaled)), 
    color = "darkgray", fill = "gray", alpha = 0.3, 
    orientation = "y", inherit.aes = FALSE) +
  scale_x_continuous(labels = function(x) ifelse(x < 0, "", x)) +
  facet_grid(name ~ `Initial Dx`, scales = 'free_y') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  ylab('')
```

## ComBat Harmonization of tau PET (tracers)

```{r tau-PET-ComBat}
#| code-fold: true
#| fig-cap: "ComBat transformed versus raw Tau PET data by tracer."

ggplot(dd %>% filter(!is.na(Tau_PET_ComBat)), 
  aes(x = Tau_PET, y = Tau_PET_ComBat, color = Tau_TRACER)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed')
```

# Publishing with NACC Data

See the [author checklist](https://www.naccdata.org/publish-with-nacc-data) for more information. If you use the `NACCADRC` R data package, please also cite [@donohue2026alzheimer].

## Funding

Work on this `R` package was funded by CLARiTI (NIH U01 AG082350). The NACC database is funded by NIA/NIH Grant U24 AG072122. SCAN was funded by NIA/NIH U24 AG067418. See [naccdata.org](https://www.naccdata.org/publish-with-nacc-data) for more information.

# References
