---
title: "NACCADRC Basic Summaries"
format: 
   html:
     toc: true
     toc_depth: 3
     code_fold: true
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{NACCADRC Basic Summaries}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 4, 
  out.width = "100%",
  class.source = "fold-hide",
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  # invalidate cached chunks when the shared data/plotting code changes
  cache.extra = tools::md5sum("_basic-summaries-data.qmd")
)

options(digits = 2, width = 150)

# Package Versions
pkgName <- "NACCADRC"
pkgFilename <- paste0('NACCADRC_', packageVersion('NACCADRC'), '.tar.gz')
```

# Merge key imaging and PHC cognitive data

{{< include _basic-summaries-data.qmd >}}

# Summarize data collection

## CLARiTI imaging distributions

```{r fig-clariti-imaging}
#| code-fold: true
#| fig-cap: "Distribution of imaging summaries among CLARiTI participants with known UDS cognitive status, amyloid PET (CL), tau PET (SUVR), and hippocampal volume (mm$^3$). Tau PET is shown as SUVR harmonized across tracers by ComBat."

plot_clariti_imaging(dd_cross)
```

## Bar charts

```{r fig-clariti-barplots}
#| code-fold: true
#| fig-cap: "Number of CLARiTI pariticpants by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(Cohort == 'CLARiTI'), by_cohort = FALSE)
```

```{r fig-all-imaging-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with any imaging data including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(Cohort != 'UDS only'))
```

```{r fig-amy-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with amyloid PET, including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(!is.na(CENTILOIDS)))
```

```{r fig-tau-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with tau PET, including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(!is.na(Tau_PET)))
```

## UpSet plots

```{r fig-clariti-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of CLARiTI participants."

# Output list CLARiTI participants with missing data for QC ----
if(FALSE){
  qc_dir <- file.path('..', 'reports', 'qc')
  dir.create(qc_dir, recursive = TRUE)
  dd_upset %>% 
    rowwise() %>%
    filter(`CLARiTI EDC` == 1, sum(c_across(MRI:`Tau PET`)) == 0) %>%
    write.csv(file.path(qc_dir, "clariti_edc_no_imaging_participants.csv"), row.names = FALSE)
  
  dd_upset %>% 
    rowwise() %>%
    filter(`CLARiTI EDC` == 1, sum(c_across(Diagnosis:`Tau PET`)) == 0) %>%
    write.csv(file.path(qc_dir, "clariti_edc_only_participants.csv"), row.names = FALSE)
  
  dd_upset %>% 
    filter(Cohort == 'CLARiTI') %>%
    write.csv(file.path(qc_dir, "clariti_participants.csv"), row.names = FALSE)
}

plot_upset(subset(dd_upset, Cohort == 'CLARiTI'))
```

```{r fig-clariti-scan-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of CLARiTI and SCAN participants."

plot_upset(subset(dd_upset, Cohort %in% c('CLARiTI', 'SCAN')))
```

```{r fig-scan-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of SCAN participants (excluding CLARiTI)."

plot_upset(subset(dd_upset, Cohort == 'SCAN'))
```

```{r fig-any-imaging-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of participants with any imaging."

plot_upset(subset(dd_upset, Cohort %in% c('CLARiTI', 'SCAN', 'Mixed protocol')))
```

```{r fig-all-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of all participants."

plot_upset(dd_upset)
```

## Scan counts

```{r tbl-scan-counts, results = 'asis'}
#| code-fold: true

tmp <- dd %>% 
  select(NACCID, CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET` = CENTILOIDS, `Volumetric MRI` = HIPPOCAMPUS, 
    `Tau PET` = Tau_PET_ComBat) %>%
  group_by(NACCID) %>%
  pivot_longer(`Amyloid PET`:`Tau PET`) %>%
  filter(!is.na(value)) %>%
  mutate(Type = factor(name, levels = c(
    "Amyloid PET", "Tau PET", "Volumetric MRI"))) %>%
  group_by(Type, NACCID) %>%
  summarise(`Serial scans` = n())

with(tmp, table(Type, `Serial scans`)) %>% 
  knitr::kable(caption = "Number of individuals who have received the given number serial scans for each scan type.")
```

## Cummulative scans by time from first scan

```{r fig-cummulative-scans}
#| code-fold: true
#| fig-cap: "Cumulative scans by time from first scan of each scan type."

dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET` = CENTILOIDS, `Volumetric MRI` = HIPPOCAMPUS, 
    `Tau PET` = Tau_PET_ComBat) %>%
  pivot_longer(`Amyloid PET`:`Tau PET`) %>%
  filter(!is.na(value), !is.na(Age), !is.na(NACCUDSD)) %>%
  mutate(Type = factor(name, levels = c(
    "Amyloid PET", "Tau PET", "Volumetric MRI"))) %>%
  group_by(NACCID, Type) %>%
  mutate(
    Age0 = min(Age, na.rm = TRUE),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down") %>%
  group_by(Type, Years) %>%
  summarise(count = n()) %>%
  mutate(`Cumulative count` = cumsum(count)) %>%
ggplot(aes(x=Years, y=`Cumulative count`, color = Type)) +
  geom_line() +
  xlab('Years from first scan of each type')
```

# Baseline characteristics

## CLARiTI participants

```{r tbl-clariti-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(Cohort == 'CLARiTI') %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("CLARiTI EDC", "Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(
    all_continuous() ~ c(
      "{mean} ({sd})",
      "{median} ({p25}, {p75})",
      "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of CLARiTI participants by baseline UDS diagnosis. Note CLARiTI participants are those with a NACCID in clariti_edc or any of the CLARiTI imaging summary files.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with any imaging

```{r tbl-any-imaging-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>%
    filter(Cohort != 'UDS only') %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of participants with any imaging by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## All participants

```{r tbl-all-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with hippocampal volumes

```{r tbl-mri-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(HIPPOCAMPUS)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"
  ),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", 
    "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants with MRI data by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with tau PET

```{r tbl-tau-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(Tau_PET)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants with tau PET data by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with amyloid PET

```{r tbl-amyloid-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(CENTILOIDS)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"
  ),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "row",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of NACC ADRC participants with amyloid PET data by baseline UDS diagnosis.") %>%
  modify_footnote_header(
    footnote = "Row-wise percentage; n (%)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

# Summary plots

## LOESS Trends

```{r loess, message=FALSE}
#| code-fold: true
#| fig-cap: "LOESS plots. For the bottom plot, memory scores are imputed using linear interpolation (when possible) and a linear mixed effects model (when interpolation was not possible) to allow for plotting against biomarker values. The linear mixed effects model included fixed effects for age (as a cubic polynomial) by initial diagnosis. Random effects included random intercepts and slopes for each participant."

plot_loess(dd)
```

## Spaghetti plots

```{r Spaghetti-centiloids}
#| code-fold: true
#| fig-cap: "Spaghetti plot of amyloid PET."

plot_spaghetti(dd, value = "CENTILOIDS", project = "Amyloid_PROJECT",
  ylab = 'Amyloid PET (CL)', alpha = 0.5)
```

```{r Spaghetti-tau-pet}
#| code-fold: true
#| fig-cap: "Spaghetti plot of tau PET."

plot_spaghetti(dd, value = "Tau_PET_ComBat", project = "Tau_PROJECT",
  ylab = 'Tau PET (MTL SUVR)')
```

```{r Spaghetti-hippocampus}
#| code-fold: true
#| fig-cap: "Spaghetti plot of hippocampal volumes."

plot_spaghetti(dd, value = "HIPPOCAMPUS", project = "MRI_PROJECT",
  ylab = 'Hippocampal volume')
```

```{r Spaghetti-cog}
#| code-fold: true
#| fig-cap: "Spaghetti plot of harmonized cognitive scores."

pd <- dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    PHC_MEM, PHC_EXF, PHC_LAN) %>%
  rename(Memory = PHC_MEM, `Exec. Function` = PHC_EXF, Language = PHC_LAN) %>%
  add_initial_dx() %>%
  pivot_longer(Memory:Language) %>%
  filter(!is.na(value), !is.na(NACCUDSD)) %>%
  known_initial_dx()

ggplot(pd, aes(x=Years, y=value)) +
  geom_point(aes(color = Etiology), alpha=0.1) +
  geom_density(aes(y = value, 
    x = after_stat(-scaled)), 
    color = "darkgray", fill = "gray", alpha = 0.3, 
    orientation = "y", inherit.aes = FALSE) +
  scale_x_continuous(labels = function(x) ifelse(x < 0, "", x)) +
  facet_grid(name ~ `Initial Dx`, scales = 'free_y',
    labeller = label_wrap_gen(width = 15)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(y = '', caption = dropped_dx_caption(pd))
```

## ComBat Harmonization of tau PET (tracers)

```{r tau-PET-ComBat}
#| code-fold: true
#| fig-cap: "ComBat transformed versus raw Tau PET data by tracer."

ggplot(dd %>% filter(!is.na(Tau_PET_ComBat)), 
  aes(x = Tau_PET, y = Tau_PET_ComBat, color = Tau_TRACER)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed')
```

# Publishing with NACC Data

See the [author checklist](https://www.naccdata.org/publish-with-nacc-data) for more information. If you use the `NACCADRC` R data package, please also cite [@donohue2026alzheimer].

## Funding

Work on this `R` package was funded by CLARiTI (NIH U01 AG082350). The NACC database is funded by NIA/NIH Grant U24 AG072122. SCAN was funded by NIA/NIH U24 AG067418. See [naccdata.org](https://www.naccdata.org/publish-with-nacc-data) for more information.

# References
