---
title: "NACCADRC Basic Summaries"
format: 
   html:
     toc: true
     toc_depth: 3
     code_fold: true
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{NACCADRC Basic Summaries}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 4, 
  out.width = "100%",
  class.source = "fold-hide",
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  # invalidate cached chunks when the shared data/plotting code changes
  cache.extra = tools::md5sum("_basic-summaries-data.qmd")
)

options(digits = 2, width = 150)

# Package Versions
pkgName <- "NACCADRC"
pkgFilename <- paste0('NACCADRC_', packageVersion('NACCADRC'), '.tar.gz')
```

# Merge key imaging and PHC cognitive data

{{< include _basic-summaries-data.qmd >}}

# Summarize data collection

## CLARiTI imaging distributions

```{r fig-clariti-imaging}
#| code-fold: true
#| fig-cap: "Imaging summaries of CLARiTI completers (participants with UDS diagnosis, SCAN/CLARiTI MRI, amyloid PET, and tau PET) by diagnosis at the CLARiTI visit: amyloid PET (CL), tau PET (SUVR, harmonized across tracers by ComBat), hippocampal volume (cm$^3$), and WMH volume (log$_{10}$ cm$^3$). Dashed lines: A/T/V cut-offs (25 CL, 1.40 SUVR, WMH top quartile)."

plot_clariti_imaging(dd_cross, complete_only = TRUE)
```

```{r fig-clariti-imaging-all}
#| code-fold: true
#| fig-cap: "Imaging summaries of all CLARiTI participants with each measure, by diagnosis at the CLARiTI visit (n per group under each violin). Dashed lines: A/T/V cut-offs."

plot_clariti_imaging(dd_cross)
```

## Bar charts

```{r fig-clariti-barplots}
#| code-fold: true
#| fig-cap: "Number of CLARiTI pariticpants by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(Cohort == 'CLARiTI'), by_cohort = FALSE)
```

```{r fig-all-imaging-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with any imaging data including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(Cohort != 'UDS only'))
```

```{r fig-amy-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with amyloid PET, including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(!is.na(CENTILOIDS)))
```

```{r fig-tau-barplots}
#| code-fold: true
#| fig-cap: "Number of pariticpants with tau PET, including CLARiTI, SCAN, and mixed-protocol by etiology and UDS clinical diagnosis."

plot_etiology_bars(dd_cross %>% filter(!is.na(Tau_PET)))
```

## UpSet plots

```{r fig-clariti-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of CLARiTI participants."

# Output list CLARiTI participants with missing data for QC ----
if(FALSE){
  qc_dir <- file.path('..', 'reports', 'qc')
  dir.create(qc_dir, recursive = TRUE)
  dd_upset %>% 
    rowwise() %>%
    filter(`CLARiTI EDC` == 1, sum(c_across(MRI:`Tau PET`)) == 0) %>%
    write.csv(file.path(qc_dir, "clariti_edc_no_imaging_participants.csv"), row.names = FALSE)
  
  dd_upset %>% 
    rowwise() %>%
    filter(`CLARiTI EDC` == 1, sum(c_across(Diagnosis:`Tau PET`)) == 0) %>%
    write.csv(file.path(qc_dir, "clariti_edc_only_participants.csv"), row.names = FALSE)
  
  dd_upset %>% 
    filter(Cohort == 'CLARiTI') %>%
    write.csv(file.path(qc_dir, "clariti_participants.csv"), row.names = FALSE)
}

plot_upset(subset(dd_upset, Cohort == 'CLARiTI'))
```

```{r fig-clariti-scan-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of CLARiTI and SCAN participants."

plot_upset(subset(dd_upset, Cohort %in% c('CLARiTI', 'SCAN')))
```

```{r fig-scan-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of SCAN participants (excluding CLARiTI)."

plot_upset(subset(dd_upset, Cohort == 'SCAN'))
```

```{r fig-any-imaging-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of participants with any imaging."

plot_upset(subset(dd_upset, Cohort %in% c('CLARiTI', 'SCAN', 'Mixed protocol')))
```

```{r fig-all-upset}
#| code-fold: true
#| fig-cap: "UpSet plot of all participants."

plot_upset(dd_upset)
```

## Scan counts

```{r tbl-scan-counts, results = 'asis'}
#| code-fold: true

tmp <- dd %>% 
  select(NACCID, CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET` = CENTILOIDS, `Volumetric MRI` = HIPPOCAMPUS, 
    `Tau PET` = Tau_PET_ComBat) %>%
  group_by(NACCID) %>%
  pivot_longer(`Amyloid PET`:`Tau PET`) %>%
  filter(!is.na(value)) %>%
  mutate(Type = factor(name, levels = c(
    "Amyloid PET", "Tau PET", "Volumetric MRI"))) %>%
  group_by(Type, NACCID) %>%
  summarise(`Serial scans` = n())

with(tmp, table(Type, `Serial scans`)) %>% 
  knitr::kable(caption = "Number of individuals who have received the given number serial scans for each scan type.")
```

## Cummulative scans by time from first scan

```{r fig-cummulative-scans}
#| code-fold: true
#| fig-cap: "Cumulative scans by time from first scan of each scan type."

dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    CENTILOIDS, HIPPOCAMPUS, Tau_PET_ComBat) %>%
  rename(
    `Amyloid PET` = CENTILOIDS, `Volumetric MRI` = HIPPOCAMPUS, 
    `Tau PET` = Tau_PET_ComBat) %>%
  pivot_longer(`Amyloid PET`:`Tau PET`) %>%
  filter(!is.na(value), !is.na(Age), !is.na(NACCUDSD)) %>%
  mutate(Type = factor(name, levels = c(
    "Amyloid PET", "Tau PET", "Volumetric MRI"))) %>%
  group_by(NACCID, Type) %>%
  mutate(
    Age0 = min(Age, na.rm = TRUE),
    `Initial Dx` = case_when(
      Age == Age0 ~ NACCUDSD,
      TRUE ~ NA),
    Years = Age - Age0) %>%
  arrange(NACCID, Years) %>%
  tidyr::fill(all_of("Initial Dx"), .direction = "down") %>%
  group_by(Type, Years) %>%
  summarise(count = n()) %>%
  mutate(`Cumulative count` = cumsum(count)) %>%
ggplot(aes(x=Years, y=`Cumulative count`, color = Type)) +
  geom_line() +
  xlab('Years from first scan of each type')
```

# Baseline characteristics

## CLARiTI participants

```{r tbl-clariti-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(Cohort == 'CLARiTI') %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("CLARiTI EDC", "Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "A_status", "Tau_PET", "Tau_PET_ComBat", "T_status", "Tau_TRACER", "HIPPOCAMPUS", "WMH", "V_status", "ATV", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(
    all_continuous() ~ c(
      "{mean} ({sd})",
      "{median} ({p25}, {p75})",
      "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "column",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = paste0("Characteristics of CLARiTI participants by UDS diagnosis at the visit closest to the CLARiTI visit (first consent date). Note CLARiTI participants are those with a NACCID in clariti_edc or any of the CLARiTI imaging summary files. Amyloid status: SCAN/CLARiTI positivity from tracer-specific thresholds on the GAAIN summary cortical SUVR (not a visual read). A/T/V: amyloid PET >= 25 CL; ComBat-harmonized meta-temporal tau PET >= 1.40 SUVR; WMH in the top quartile of SCAN/CLARiTI participants (", sprintf("%.1f", atv_cutoffs[["wmh_cc"]]), " cm3). A/T/V profile for participants with all three measures.")) %>%
  modify_footnote_header(
    footnote = "n (%) within diagnosis (column)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with any imaging

```{r tbl-any-imaging-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>%
    filter(Cohort != 'UDS only') %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "column",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of participants with any imaging by baseline UDS diagnosis (for CLARiTI participants, the UDS visit closest to the CLARiTI visit).") %>%
  modify_footnote_header(
    footnote = "n (%) within diagnosis (column)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## All participants

```{r tbl-all-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "column",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants by baseline UDS diagnosis (for CLARiTI participants, the UDS visit closest to the CLARiTI visit).") %>%
  modify_footnote_header(
    footnote = "n (%) within diagnosis (column)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with hippocampal volumes

```{r tbl-mri-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(HIPPOCAMPUS)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"
  ),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "column",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", 
    "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants with MRI data by baseline UDS diagnosis (for CLARiTI participants, the UDS visit closest to the CLARiTI visit).") %>%
  modify_footnote_header(
    footnote = "n (%) within diagnosis (column)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with tau PET

```{r tbl-tau-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(Tau_PET)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "column",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of all participants with tau PET data by baseline UDS diagnosis (for CLARiTI participants, the UDS visit closest to the CLARiTI visit).") %>%
  modify_footnote_header(
    footnote = "n (%) within diagnosis (column)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

## Participants with amyloid PET

```{r tbl-amyloid-char}
#| code-fold: true

tbl_summary(
  data = dd_cross %>% filter(!is.na(CENTILOIDS)) %>%
    mutate(across(where(is.factor), fct_drop)),
  by = NACCUDSD,
  include = c("Etiology", "Age", "SEX", "EDUC", "RACE", "HISPANIC", "AMYLOID_STATUS", "CENTILOIDS", "HIPPOCAMPUS", "Tau_PET", "Tau_TRACER", "PHC_MEM", "PHC_EXF", "PHC_LAN"),
  type = all_continuous() ~ "continuous2",
  statistic = list(all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({p25}, {p75})",
    "{min}, {max}"
  ),
    all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1,
  percent = "column",
  missing_text = "(Missing)") %>%
  add_overall(last = TRUE) %>%
  add_stat_label(label = all_continuous2() ~ c("Mean (SD)", "Median (Q1, Q3)", "Range")) %>%
  modify_caption(caption = "Characteristics of NACC ADRC participants with amyloid PET data by baseline UDS diagnosis (for CLARiTI participants, the UDS visit closest to the CLARiTI visit).") %>%
  modify_footnote_header(
    footnote = "n (%) within diagnosis (column)",
    columns = all_stat_cols(),
    replace = TRUE) %>%
  bold_labels()
```

# Summary plots

## LOESS Trends

```{r loess, message=FALSE}
#| code-fold: true
#| fig-cap: "LOESS plots. For the bottom plot, memory scores are imputed using linear interpolation (when possible) and a linear mixed effects model (when interpolation was not possible) to allow for plotting against biomarker values. The linear mixed effects model included fixed effects for age (as a cubic polynomial) by initial diagnosis. Random effects included random intercepts and slopes for each participant."

plot_loess(dd)
```

## Spaghetti plots

```{r Spaghetti-centiloids}
#| code-fold: true
#| fig-cap: "Spaghetti plot of amyloid PET."

plot_spaghetti(dd, value = "CENTILOIDS", project = "Amyloid_PROJECT",
  ylab = 'Amyloid PET (CL)', alpha = 0.5)
```

```{r Spaghetti-tau-pet}
#| code-fold: true
#| fig-cap: "Spaghetti plot of tau PET."

plot_spaghetti(dd, value = "Tau_PET_ComBat", project = "Tau_PROJECT",
  ylab = 'Tau PET (MTL SUVR)')
```

```{r Spaghetti-hippocampus}
#| code-fold: true
#| fig-cap: "Spaghetti plot of hippocampal volumes."

plot_spaghetti(dd, value = "HIPPOCAMPUS", project = "MRI_PROJECT",
  ylab = 'Hippocampal volume')
```

```{r Spaghetti-cog}
#| code-fold: true
#| fig-cap: "Spaghetti plot of harmonized cognitive scores."

pd <- dd %>% 
  select(NACCID, Age, Etiology, NACCUDSD, NACCETPR,
    PHC_MEM, PHC_EXF, PHC_LAN) %>%
  rename(Memory = PHC_MEM, `Exec. Function` = PHC_EXF, Language = PHC_LAN) %>%
  add_initial_dx() %>%
  pivot_longer(Memory:Language) %>%
  filter(!is.na(value), !is.na(NACCUDSD)) %>%
  known_initial_dx()

ggplot(pd, aes(x=Years, y=value)) +
  geom_point(aes(color = Etiology), alpha=0.1) +
  geom_density(aes(y = value, 
    x = after_stat(-scaled)), 
    color = "darkgray", fill = "gray", alpha = 0.3, 
    orientation = "y", inherit.aes = FALSE) +
  scale_x_continuous(labels = function(x) ifelse(x < 0, "", x)) +
  facet_grid(name ~ `Initial Dx`, scales = 'free_y',
    labeller = label_wrap_gen(width = 15)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(y = '', caption = dropped_dx_caption(pd))
```

## CLARiTI spaghetti plots from the CLARiTI visit

Time 0 is each CLARiTI participant's CLARiTI visit (first consent date, CNSTDT1;
screening date if not available). Scans before the CLARiTI visit, e.g. from
SCAN or legacy mixed protocol imaging, have negative times. Panels are the UDS
diagnosis at the visit closest to the CLARiTI visit.

```{r Spaghetti-clariti-centiloids}
#| code-fold: true
#| fig-cap: "Amyloid PET of CLARiTI participants by years from the CLARiTI visit."

plot_spaghetti_clariti(dd, value = "CENTILOIDS", project = "Amyloid_PROJECT",
  ylab = 'Amyloid PET (CL)', alpha = 0.5)
```

```{r Spaghetti-clariti-tau-pet}
#| code-fold: true
#| fig-cap: "Tau PET of CLARiTI participants by years from the CLARiTI visit."

plot_spaghetti_clariti(dd, value = "Tau_PET_ComBat", project = "Tau_PROJECT",
  ylab = 'Tau PET (MTL SUVR)')
```

```{r Spaghetti-clariti-hippocampus}
#| code-fold: true
#| fig-cap: "Hippocampal volumes of CLARiTI participants by years from the CLARiTI visit."

plot_spaghetti_clariti(dd, value = "HIPPOCAMPUS", project = "MRI_PROJECT",
  ylab = 'Hippocampal volume (cm3)')
```

## ComBat Harmonization of tau PET (tracers)

```{r tau-PET-ComBat}
#| code-fold: true
#| fig-cap: "ComBat transformed versus raw Tau PET data by tracer."

ggplot(dd %>% filter(!is.na(Tau_PET_ComBat)), 
  aes(x = Tau_PET, y = Tau_PET_ComBat, color = Tau_TRACER)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed')
```

# Publishing with NACC Data

See the [author checklist](https://www.naccdata.org/publish-with-nacc-data) for more information. If you use the `NACCADRC` R data package, please also cite [@donohue2026alzheimer].

## Funding

Work on this `R` package was funded by CLARiTI (NIH U01 AG082350). The NACC database is funded by NIA/NIH Grant U24 AG072122. SCAN was funded by NIA/NIH U24 AG067418. See [naccdata.org](https://www.naccdata.org/publish-with-nacc-data) for more information.

# References
