---
title: "Latent-Class, Latent-Time Joint Models for CLARiTI: Candidate Model Comparison"
subtitle: "Model space, fitting workflow, WAIC/LOOIC comparison, and summary of the best-fitting model"
author: "Li, Donohue, Langford, Insel, et al."
date: today
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

<!--
Latent-class mixed models for five ATN and cognitive outcomes in NACC
(including CLARiTI), fit with lcmm::mpjlcmm(): one latent process per outcome,
linked through a single shared latent-class variable. Data are built by
build-data.R from the shared NACCADRC data preparation
(vignettes/_basic-summaries-data.qmd). Model fits are cached under fits/ and
are run by the analyst, not as part of drafting this document.
-->

```{r}
#| label: packages
# lcmm (hlme + mpjlcmm) fits the models; splines builds the age basis.
# cmdstanr/posterior/bayesplot/loo were only needed for the abandoned
# Bayesian Stan grid (@sec-limitations).
library(lcmm)
library(splines)
library(tidyverse)
library(patchwork)
library(knitr)
library(kableExtra)
library(data.table)
library(gtsummary)

set.seed(20260910)

options(htmltools.dir.version = FALSE, knitr.kable.NA = "")

knitr::opts_chunk$set(
  echo = FALSE,
  tidy = FALSE,
  collapse = TRUE,
  fig.path = "",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  out.extra = "",
  out.width = "\\linewidth",
  crop = TRUE,
  fig.pos = "htbp",
  cache = FALSE,
  comment = "",
  fig.dpi = 300
)

dir.create("fits", showWarnings = FALSE, recursive = TRUE)
```

```{r}
#| label: ggplot-setup
theme_jama <- function(base_size = 10) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      plot.title = element_text(size = base_size + 1, face = "bold", hjust = 0, margin = margin(b = 10)),
      plot.subtitle = element_text(size = base_size, hjust = 0, margin = margin(b = 10)),
      plot.caption = element_text(size = base_size - 1, hjust = 1, margin = margin(t = 10)),
      axis.title = element_text(size = base_size, face = "bold"),
      axis.text = element_text(size = base_size - 1, color = "black"),
      axis.line = element_line(color = "black", linewidth = 0.5),
      axis.ticks = element_line(color = "black", linewidth = 0.5),
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.title = element_text(size = base_size, face = "bold"),
      legend.text = element_text(size = base_size - 1),
      legend.key = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.position = "bottom",
      strip.background = element_blank(),
      strip.text = element_text(size = base_size - 2, face = "bold", vjust = 1, color = "black"),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

jama_colors <- c("#374E55", "#DF8F44", "#00A1D5", "#B24745", "#79AF97", "#6A6599", "#80796B")
scale_color_jama <- function(...) scale_color_manual(values = jama_colors, ...)
scale_fill_jama  <- function(...) scale_fill_manual(values = jama_colors, ...)
```

```{r}
#| label: ecdf-functions
ecdf.func <- function(x, weights = NULL,
  type = c("i/n", "(i-1)/(n-1)", "i/(n+1)"),
  normwt = TRUE, na.rm = TRUE,
  method = "linear", rule = 2, ...) {
  type <- match.arg(type)
  ecdf.data <- Hmisc::wtd.Ecdf(x, weights = weights, type = type,
    normwt = normwt, na.rm = na.rm)
  with(ecdf.data, approxfun(x, ecdf, method = method, na.rm = na.rm, rule = rule, ties = mean, ...))
}
```

# Introduction

An earlier exploratory analysis (`lctjm-clariti-report.Rmd`) fit a sequence
of hand-tuned latent-time and latent-class joint models to five harmonized
outcomes (hippocampal atrophy, amyloid PET, tau PET, memory, and executive
function) and found evidence, via LOOIC, for three latent classes with
class-specific intercepts. A systematic follow-up built two general-purpose
Stan programs (`ltjm_general.stan`/`lctjm_general.stan`) around a shared
participant-specific *latent time shift* $\delta_i$; every variant converged
at $K \le 2$ but none reached acceptable convergence at $K = 3$
(@sec-limitations).

A second attempt used `lcmm::multlcmm()`. `multlcmm` treats all outcomes as
measurements of a *single* latent process: only that shared process can have
class-specific trajectories, and `contrast()` terms give outcome-specific
deviations that are common to every class. The intended model -- each
outcome with its own class-specific trajectory -- was therefore not one
`multlcmm` can express, and that specification did not converge.

This document fits that intended model with `lcmm::mpjlcmm()`: **one latent
process per outcome, linked through a single shared latent-class
variable**. Each participant belongs to one class, and that class determines
the participant's mean trajectory over age on every outcome; each outcome
has its own class-specific age spline and its own random effects. There is
no latent time shift: chronological age is the time scale.

If this converges, the next step is to reintroduce a participant-specific
latent time shift, using the `mpjlcmm` fit -- class-specific trajectory
shapes, random-effect variances, residual variances, and class proportions
-- to inform the priors and parameterization of the latent-time model
(@sec-next-steps).

Data come from all NACC participants with at least one of amyloid PET, tau
PET, or hippocampal volume, plus memory (the "ATN union" cohort), of whom
CLARiTI participants are a subgroup. The document sweeps the number of
latent classes ($K = 1, \dots, 4$), compares candidates using BIC and
classification quality (mean posterior assignment probability, entropy), and
summarizes the best-fitting model. Model fitting is run interactively (fits
are cached in `fits/`); re-render afterwards for the comparison and summary
sections.

# Data {#sec-data}

## Long-format data assembly

The long-format participant-visit data are built by `build-data.R` (run
from this directory: `Rscript build-data.R`), which reuses the NACCADRC
package's shared data preparation (`vignettes/_basic-summaries-data.qmd`):

- **Hippocampal atrophy:** hippocampal volume from SCAN/CLARiTI MRI
  (`HIPPOCAMPUS`) and legacy mixed protocol MRI (`HIPPOVOL`), with
  implausible volumes (< 1.5 cm$^3$) removed; adjusted for intracranial
  volume (`ESTIMATEDTOTALINTRACRANIALVOL`/1000 or `NACCICV`, cm$^3$),
  Z-scored, and sign flipped so higher values mean more atrophy.
- **Amyloid PET:** centiloids, latest processing per scan, SCAN, CLARiTI, and
  legacy mixed protocol.
- **Tau PET:** meta-temporal SUVR harmonized across tracers by ComBat.
- **Cognition:** PHC harmonized memory, executive function, and language
  composites dated by UDS visit, sign flipped so higher values are worse.
- **Covariates:** sex, education, APOE genotype, and first UDS diagnosis;
  participants missing any covariate are excluded.

```{r}
#| label: load-data
data(clariti_edc, package = "NACCADRC")
clariti_ids <- unique(clariti_edc$NACCID)

ddl_atn_union <- readRDS('clariti-long-atn-union-data.RDS')
ddl <- ddl_atn_union %>%
  # e3e3 (the most common genotype) as the reference for class membership
  mutate(APOE = forcats::fct_relevel(APOE, "e3e3"))
```

## Standardizing outcomes to a common severity scale

Because the five outcomes are on incommensurable scales (Z-scored
hippocampal atrophy, amyloid centiloids, tau SUVR, and harmonized cognitive
composites), each is mapped to a common severity metric before joint
modeling: a diagnosis-reweighted empirical CDF converts each raw outcome to
a severity quantile $Q \in (0,1)$ (reweighted so that Cognitively
Normal/MCI/Dementia are represented as in the general population rather
than in this enriched sample), and $Z = \Phi^{-1}(Q)$ places all outcomes
on a common standard-normal severity scale. This transform, and the
outcome coding, are unchanged from the original draft.

```{r}
#| label: clariti-q-z-scores
ddl_ecdf <- ddl %>%
  mutate(
    # behavioral impairment only (UDS v4) has no cognitive impairment, so it
    # is weighted with normal cognition rather than as its own (tiny) group
    Dx_ecdf = forcats::fct_collapse(Dx,
      `Normal cognition` = intersect(c("Normal cognition",
        "Behavioral impairment only"), levels(Dx)),
      MCI = c("MCI", "Impaired-not-MCI"))
  )

ecdf_maps_collapsed <- ddl_ecdf %>%
  dplyr::count(Outcome, Dx_ecdf, name = "n_dx") %>%
  group_by(Outcome) %>%
  mutate(prop_dx = n_dx / sum(n_dx)) %>%
  ungroup() %>%
  right_join(ddl_ecdf, by = c("Outcome", "Dx_ecdf")) %>%
  mutate(weight = 1 / prop_dx) %>%
  group_by(Outcome) %>%
  summarise(
    Y_to_Q = list(ecdf.func(x = Y, weights = weight, type = "i/(n+1)")),
    .groups = "drop"
  )

ddl_qz <- ddl_ecdf %>%
  left_join(ecdf_maps_collapsed, by = "Outcome") %>%
  mutate(
    Q = map2_dbl(Y_to_Q, Y, ~ .x(.y)),
    Z = qnorm(Q)
  ) %>%
  arrange(id, outcome)
```

## Outcome set and coding

As in the original draft, Language is dropped, leaving $J = 5$ outcomes.

```{r}
#| label: outcome-coding
recode_outcome <- function(dat) {
  dat %>%
    filter(Outcome != "Language") %>%
    mutate(
      outcome = case_when(
        Outcome == "Hipp. atrophy (Z-score)" ~ 1,
        Outcome == "Amyloid PET (CL)"        ~ 2,
        Outcome == "Tau PET (SUVR)"          ~ 3,
        Outcome == "Memory"                  ~ 4,
        Outcome == "Exec Function"           ~ 5,
        TRUE ~ NA_real_
      )
    ) %>%
    arrange(id, outcome, years)
}

ddl <- recode_outcome(ddl)
ddl_qz <- recode_outcome(ddl_qz)

J <- length(unique(ddl_qz$outcome))

outcome_map <- ddl_qz %>% distinct(outcome, Outcome) %>% arrange(outcome)
```

# Candidate model space {#sec-grid}

The number of latent classes, $K = 1, \dots, 4$, is the only axis compared.
The per-outcome sub-model structure (@sec-methods-model) is fixed in advance
from how much longitudinal data each outcome has (@tbl-outcome-coverage).

```{r}
#| label: spec-grid
# Search settings per K. K = 2 and 3 were fit with one 50-start search;
# K = 4 uses two independent 15-start searches whose agreement is checked
# (@sec-stage1). K = 1 has a single solution and needs no search.
spec_grid <- tibble::tibble(
  K        = 1:4,
  n_starts = c(NA, 50, 50, 15),
  n_runs   = c(1, 1, 1, 2)) %>%
  mutate(model_id = paste0("mpjlcmm_K", K))
```

```{r}
#| label: tab-spec-grid
kable(spec_grid, caption = "Candidate specification grid: number of latent classes.")
```

# Statistical methods {#sec-methods}

## Model {#sec-methods-model}

For participant $i$, outcome $j = 1, \dots, J$, and observation $v$ at
age $a_{ijv}$, outcome $j$'s severity score (@sec-data) follows its own
linear mixed model,

$$
Z_{ijv} = f_{jk(i)}(a_{ijv}) + \mathbf{x}_i^{\mathsf T}\boldsymbol\beta_j
  + b_{0ij} + b_{1ij}\, a_{ijv} + \varepsilon_{ijv}, \qquad
  \varepsilon_{ijv} \sim \operatorname{Normal}(0, \sigma_j^2),
$$

where $k(i) \in \{1, \dots, K\}$ is participant $i$'s latent class,
**shared across all $J$ outcomes**; $f_{jk}$ is a class- and
outcome-specific trajectory over age (class-specific intercept plus a
natural cubic spline); $\mathbf x_i$ contains sex and education with
outcome-specific effects; and $(b_{0ij}, b_{1ij})$ are outcome-specific
random effects. The latent class is the only link between outcomes: given
the class, the $J$ outcome processes are independent. This is
`lcmm::mpjlcmm()` with one `hlme()` sub-model per outcome.

Each outcome's sub-model is matched to its data (@tbl-outcome-coverage):
amyloid and tau PET are mostly single scans (about 85% of participants), so
they get a random intercept only and a 2-df spline; hippocampal atrophy,
memory, and executive function have repeated measures for many participants
and get a random intercept and slope on age and a 3-df spline.

## Class membership model

Class membership follows a multinomial logistic regression on APOE genotype
(`classmb = ~ APOE`, reference genotype e3e3), with class 1 as the reference
class.

## Local optima and multiple starting values {#sec-methods-optima}

The likelihood of a $K$-class mixture is multimodal, so a single
optimization can stop at a local maximum. For each $K \ge 2$,
`lcmm::gridsearch()` runs a short optimization from many random
perturbations of the $K = 1$ joint fit and fully optimizes the best start.

## Estimation

Parameters are estimated by maximum likelihood with a
Marquardt-Levenberg algorithm. Uncertainty is reported via Wald standard
errors from the Hessian; `predictY(..., draws = TRUE)` gives confidence
bands for fitted trajectories by Monte Carlo draws from the asymptotic
distribution of the estimates. A fit is accepted only with convergence code
1 (converged with a positive-definite Hessian).

## Model comparison

Candidate $K$ are compared by BIC ($-2\times\text{log-likelihood} + p\log N$,
$p$ parameters, $N$ participants; lower is better), alongside mean posterior
class-assignment probability and normalized entropy of the posterior
class-membership probabilities. A lower BIC with posterior probabilities
near $1/K$ for many participants improves fit without separating
participants into classes.

# Model fitting {#sec-fitting}

## Helper functions

```{r}
#| label: helper-wide-data
# One row per (id, age); each outcome's severity Z-score in its own column
# (NA where not measured at that age). Each hlme() sub-model uses only the
# rows where its outcome is observed.
build_wide_data <- function(ddl_qz) {
  ddl_qz %>%
    transmute(id, outcome, age, SEX, APOE, EDUC, Z,
      outcome_lab = paste0("Z", outcome)) %>%
    tidyr::pivot_wider(
      id_cols = c(id, age, SEX, APOE, EDUC),
      names_from = outcome_lab,
      values_from = Z,
      values_fn = ~ mean(.x, na.rm = TRUE)
    ) %>%
    arrange(id, age)
}

wide_qz <- build_wide_data(ddl_qz)
stopifnot(nrow(wide_qz) == nrow(distinct(wide_qz, id, age)))

# age_c: age centered at the sample mean, for optimizer conditioning
age_center <- mean(wide_qz$age, na.rm = TRUE)
wide_qz <- wide_qz %>% mutate(age_c = age - age_center)

# Natural spline bases computed once and stored as columns. Calling ns()
# inside a model formula recomputes the knots on any newdata passed to
# predictY(), giving wrong predicted trajectories; these basis objects are
# reused for prediction instead (add_age_basis()).
age_basis <- list(
  ns2 = splines::ns(wide_qz$age_c, df = 2),
  ns3 = splines::ns(wide_qz$age_c, df = 3))

add_age_basis <- function(data) {
  for (nm in names(age_basis)) {
    b <- predict(age_basis[[nm]], data$age_c)
    # plain numeric matrix: the "ns"/"basis" class would otherwise carry into
    # each column, which hlme() cannot handle
    b <- matrix(as.numeric(b), nrow = nrow(b),
      dimnames = list(NULL, paste0(nm, "_", seq_len(ncol(b)))))
    data <- bind_cols(data %>% select(-any_of(colnames(b))), as_tibble(b))
  }
  data
}
wide_qz <- add_age_basis(wide_qz)
```

```{r}
#| label: tbl-outcome-coverage
#| tbl-cap: "Per-outcome data: participants, observations, and how many participants have repeated measures."
outcome_coverage <- ddl_qz %>%
  filter(!is.na(Z)) %>%
  count(outcome, Outcome, id, name = "n_obs_id") %>%
  group_by(outcome, Outcome) %>%
  summarise(
    n_subjects = n(),
    n_obs = sum(n_obs_id),
    pct_2plus = round(100 * mean(n_obs_id >= 2), 1),
    pct_3plus = round(100 * mean(n_obs_id >= 3), 1),
    .groups = "drop") %>%
  left_join(
    ddl_qz %>% filter(!is.na(Z)) %>% group_by(outcome) %>%
      summarise(age_min = min(age), age_max = max(age)),
    by = "outcome")

kable(outcome_coverage, digits = 1)
```

```{r}
#| label: helper-mpjlcmm-fit
# One hlme() sub-model per outcome (@sec-methods-model). spline: which age
# basis; random: random-effects formula.
outcome_specs <- tibble::tribble(
  ~outcome, ~spline, ~random,
  1,        "ns3",   "~ 1 + age_c",   # hippocampal atrophy
  2,        "ns2",   "~ 1",           # amyloid PET: mostly single scans
  3,        "ns2",   "~ 1",           # tau PET: mostly single scans
  4,        "ns3",   "~ 1 + age_c",   # memory
  5,        "ns3",   "~ 1 + age_c"    # executive function
) %>%
  mutate(
    y = paste0("Z", outcome),
    basis_terms = map_chr(spline, ~ paste(
      grep(paste0("^", .x, "_"), names(wide_qz), value = TRUE), collapse = " + ")),
    fixed = paste(y, "~", basis_terms, "+ SEX + EDUC"),
    mixture = paste("~", basis_terms))

# hlme() sub-model for one outcome. K = 1: fitted. K >= 2: an unfitted
# template (maxiter = 0) started from the K = 1 fit (m1_outcome), as
# mpjlcmm() expects. The formulas and m1_outcome are spliced into the call
# with bquote() because mpjlcmm() and update() re-evaluate the sub-models'
# stored calls in the global environment, and hlme() reads B = random(...)
# unevaluated (so do.call() cannot be used).
outcome_model <- function(spec, K, m1_outcome = NULL) {
  fixed <- as.formula(spec$fixed)
  mixture <- as.formula(spec$mixture)
  random_f <- as.formula(spec$random)
  if (K == 1) {
    fit <- eval(bquote(hlme(fixed = .(fixed), random = .(random_f),
      subject = "id", data = wide_qz, maxiter = 500, verbose = FALSE)))
    check_converged(fit, paste0(spec$y, "_K1"))
  } else {
    eval(bquote(hlme(fixed = .(fixed), mixture = .(mixture),
      random = .(random_f), subject = "id", ng = .(K), data = wide_qz,
      B = random(.(m1_outcome)), maxiter = 0, verbose = FALSE)))
  }
}

# K = 1 sub-models (fitted once and cached) and K-class templates
load_m1 <- function() {
  map(seq_len(nrow(outcome_specs)), function(r) {
    spec <- outcome_specs[r, ]
    fit_or_load(file.path("fits", paste0("hlme_", spec$y, "_K1.RDS")),
      outcome_model(spec, K = 1))
  })
}
# Each K-class template is assigned to its own global name (template_K2_Z1,
# ...): with cl, gridsearch() exports the mpjlcmm() sub-models to the worker
# processes by name, so longitudinal must be list(<names>), not one list
# object. Returns the names.
make_templates <- function(K, m1) {
  nm <- paste0("template_K", K, "_", outcome_specs$y)
  for (j in seq_along(nm)) {
    assign(nm[j],
      outcome_model(outcome_specs[j, ], K = K, m1_outcome = m1[[j]]),
      envir = globalenv())
  }
  nm
}

# conv codes (see ?hlme): 1 = converged; 2 = maximum iterations reached;
# 3 = converged but Hessian not positive definite; 4/5 = numerical problems.
# A non-converged fit is saved to unconverged_path (if given) so long runs
# are not lost: it can be inspected, or used as a starting point (B =).
check_converged <- function(fit, model_id, unconverged_path = NULL) {
  if (fit$conv != 1) {
    if (!is.null(unconverged_path)) saveRDS(fit, unconverged_path)
    stop(sprintf(paste(
      "%s did not converge (conv = %d).%s",
      "Try more iterations, a simpler random-effects structure for the",
      "sparsest outcomes, or fewer spline df."), model_id, fit$conv,
      if (is.null(unconverged_path)) "" else
        paste0(" Saved for inspection to ", unconverged_path, ".")))
  }
  fit
}

fit_or_load <- function(path, expr) {
  if (file.exists(path)) return(readRDS(path))
  fit <- expr
  saveRDS(fit, path)
  fit
}

# One gridsearch() for K >= 2: random starts perturbing the K = 1 joint
# fit (j1), short searches on n_cores workers, then a full optimization of
# the best start.
run_gridsearch <- function(K, n_starts, n_cores, m1, j1) {
  longitudinal <- as.call(c(as.name("list"),
    lapply(make_templates(K, m1), as.name)))
  eval(bquote(gridsearch(rep = .(n_starts), maxiter = 30, minit = j1,
    cl = .(n_cores),
    mpjlcmm(longitudinal = .(longitudinal), subject = "id", ng = .(K),
      # verbose = TRUE shows the final estimation's iterations; the search
      # runs on the workers print nowhere (or to the console if serial)
      classmb = ~ APOE, data = wide_qz, maxiter = 500, verbose = TRUE))))
}

# n_runs > 1: repeat the search n_runs times (each cached as
# <model_id>_run<r>.RDS) and keep the run with the highest log-likelihood.
# Agreement between runs is checked in run_agreement(). With parallel
# workers, each run draws different random starts regardless of seed.
# n_cores: worker processes for gridsearch(); NULL runs serially.
fit_or_load_mpjlcmm <- function(model_id, K, n_starts = 50, n_runs = 1,
  seed = 20260101, n_cores = max(1, parallel::detectCores() - 1)) {
  path <- file.path("fits", paste0(model_id, ".RDS"))
  if (file.exists(path)) return(readRDS(path))
  set.seed(seed)

  m1 <- load_m1()

  if (K == 1) {
    fit <- mpjlcmm(longitudinal = m1, subject = "id", ng = 1,
      data = wide_qz, verbose = FALSE)
    check_converged(fit, model_id)
  } else {
    j1_path <- file.path("fits", "mpjlcmm_K1.RDS")
    stopifnot("Fit and cache K = 1 first; K >= 2 starts from it." =
      file.exists(j1_path))
    j1 <- readRDS(j1_path)

    run_paths <- if (n_runs == 1) path else
      file.path("fits", paste0(model_id, "_run", seq_len(n_runs), ".RDS"))
    runs <- purrr::map(seq_along(run_paths), function(r) {
      if (file.exists(run_paths[r])) return(readRDS(run_paths[r]))
      set.seed(seed + r)
      fit_r <- run_gridsearch(K, n_starts, n_cores, m1, j1)
      check_converged(fit_r, basename(run_paths[r]),
        unconverged_path = sub("\\.RDS$", "_unconverged.RDS", run_paths[r]))
      saveRDS(fit_r, run_paths[r])
      fit_r
    })
    fit <- runs[[which.max(purrr::map_dbl(runs, "loglik"))]]
  }
  saveRDS(fit, path)
  fit
}

# Agreement between repeated searches for one K: log-likelihood, class
# proportions (sorted, since class labels are arbitrary), and the share of
# participants assigned to the same class after the best relabeling.
run_agreement <- function(model_id, K, n_runs) {
  run_paths <- file.path("fits", paste0(model_id, "_run", seq_len(n_runs), ".RDS"))
  if (!all(file.exists(run_paths))) return(NULL)
  runs <- purrr::map(run_paths, readRDS)
  class_of <- purrr::map(runs, ~ setNames(.x$pprob$class, .x$pprob[[1]]))
  ids <- Reduce(intersect, purrr::map(class_of, names))
  perms <- function(v) if (length(v) <= 1) list(v) else
    do.call(c, lapply(seq_along(v), function(i)
      lapply(perms(v[-i]), function(p) c(v[i], p))))
  best_match <- function(a, b) max(purrr::map_dbl(perms(seq_len(K)),
    ~ mean(.x[a] == b)))
  tibble(
    run = seq_len(n_runs),
    loglik = purrr::map_dbl(runs, "loglik"),
    loglik_diff_from_best = loglik - max(loglik),
    class_proportions = purrr::map_chr(runs, ~ paste(
      sprintf("%.3f", sort(prop.table(table(factor(.x$pprob$class,
        levels = seq_len(K)))), decreasing = TRUE)), collapse = " / ")),
    pct_same_class_as_run1 = purrr::map_dbl(class_of, ~ round(100 *
      best_match(class_of[[1]][ids], .x[ids]), 1)))
}
```

## Stage 1: screen $K$ {#sec-stage1}

```{r}
#| label: stage1-fit
dir.create("fits", showWarnings = FALSE)
stage1_fits <- purrr::pmap(spec_grid, function(K, n_starts, n_runs, model_id) {
  fit_or_load_mpjlcmm(model_id = model_id, K = K,
    n_starts = n_starts, n_runs = n_runs)
})
names(stage1_fits) <- paste0("K", spec_grid$K)
```

```{r}
#| label: tbl-run-agreement
#| tbl-cap: "Agreement between repeated searches (K fit with more than one run). Runs agreeing in log-likelihood (within about 0.1) and class assignment suggest the best solution was found."
run_agreement_table <- spec_grid %>%
  filter(n_runs > 1) %>%
  purrr::pmap_dfr(function(K, n_starts, n_runs, model_id) {
    agreement <- run_agreement(model_id, K, n_runs)
    if (!is.null(agreement)) bind_cols(K = K, n_starts = n_starts, agreement)
  })
kable(run_agreement_table, digits = 2)
```

```{r}
#| label: stage1-comparison
diagnostics_row <- function(fit, K) {
  npm <- length(fit$best)
  N <- fit$ns  # participants contributing to the fit
  class_diag <- if (K == 1) {
    tibble(entropy = NA_real_, mean_postprob = NA_real_)
  } else {
    # fit$pprob: subject id, modal class, then probYT1..probYTK
    probs <- as.matrix(fit$pprob[, grep("^probYT", names(fit$pprob))])
    entropy <- 1 - mean(rowSums(-probs * log(probs + 1e-12)) / log(K))
    tibble(entropy = entropy, mean_postprob = mean(apply(probs, 1, max)))
  }
  tibble(K = K, loglik = fit$loglik, npm = npm, N = N, conv = fit$conv,
    BIC = -2 * fit$loglik + npm * log(N)) %>%
    bind_cols(class_diag)
}

diagnostics_summary <- purrr::imap_dfr(stage1_fits, function(fit, nm) {
  bind_cols(model_id = nm, diagnostics_row(fit, as.integer(sub("K", "", nm))))
})
```

```{r}
#| label: tbl-diagnostics-summary
kable(
  diagnostics_summary,
  digits = 3,
  caption = paste(
    "Convergence and classification-quality diagnostics across K.",
    "Lower BIC is better; entropy and mean posterior probability closer",
    "to 1 indicate cleaner class separation."
  )
)
```

```{r}
#| label: converged-models
converged_ids <- purrr::keep(stage1_fits, ~ .x$conv == 1) %>% names()
diagnostics_summary_converged <- diagnostics_summary %>% filter(model_id %in% converged_ids)
```

# Model comparison {#sec-comparison}

```{r}
#| label: bic-comparison
comparison_metrics <- diagnostics_summary_converged %>%
  arrange(BIC)
```

```{r}
#| label: tbl-comparison-overall
#| tbl-cap: "BIC and classification-quality diagnostics for all converged candidate models, sorted by BIC (lower is better)."
comparison_metrics %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kableExtra::kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

BIC alone can favor a larger $K$ that improves fit without separating
participants into distinct classes (@sec-methods). Before accepting the
BIC-minimizing $K$, check that its entropy and mean posterior
class-assignment probability are also reasonable: entropy well below 0.6,
or a mean assignment probability not much above $1/K$, suggests the BIC
improvement is not coming from well-separated classes.

## Selecting the best-fitting model

```{r}
#| label: best-model
best_row <- comparison_metrics %>%
  slice_min(BIC, n = 1)

K_final <- as.integer(sub("^K", "", best_row$model_id))
fit_final <- stage1_fits[[paste0("K", K_final)]]

best_row
```

# Summary of the best-fitting model {#sec-summary}

```{r}
#| label: prep-final-fit
summary(fit_final)
```

## Parameter summary

```{r}
#| label: tbl-posterior-summary
#| tbl-cap: "Parameter estimates for the selected model (Wald SE from the Hessian)."
est <- fit_final$best
se <- sqrt(diag(vcov(fit_final)))
param_table <- tibble(parameter = names(est), estimate = est, se = se,
  wald_p = 2 * pnorm(-abs(est / se)))

kable(param_table, digits = 3, booktabs = TRUE, longtable = TRUE)
```

## Class assignment (K \(\geq\) 2 models only)

```{r}
#| label: extract-class-assignments
class_lookup <- if (K_final > 1) {
  as_tibble(fit_final$pprob) %>%
    rename(id = 1, latent_class_num = class) %>%
    mutate(latent_class = factor(latent_class_num, levels = seq_len(K_final),
      labels = paste("Class", seq_len(K_final))))
} else {
  tibble(id = sort(unique(wide_qz$id)), latent_class = factor("Class 1"))
}

# Small id -> class lookup, e.g. for refitting a latent-time model within
# each class later.
saveRDS(class_lookup, file.path("fits", "class_lookup_final.RDS"))
```

## Baseline characteristics by class

```{r}
#| label: tbl-baseline-by-class
#| tbl-cap: "Characteristics of participants by modal latent class (best-fitting model)."
dd_subject <- ddl %>%
  select(id, NACCID, Dx, age0, RACE, SEX, EDUC, HISPANIC, APOE) %>%
  distinct() %>%
  mutate(CLARiTI = if_else(NACCID %in% clariti_ids, "Yes", "No"))

dd_subject_class <- dd_subject %>%
  left_join(class_lookup %>% select(id, latent_class), by = "id", relationship = "one-to-one") %>%
  left_join(
    ddl %>%
      filter(!is.na(Y), !is.na(years)) %>%
      group_by(id, NACCID, Outcome) %>%
      slice_min(order_by = years, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(id, NACCID, Outcome, Y) %>%
      pivot_wider(names_from = Outcome, values_from = Y),
    by = c("id", "NACCID"), relationship = "one-to-one"
  )

gtsummary::tbl_summary(
  data = dd_subject_class,
  by = latent_class,
  include = c(Dx, age0, SEX, EDUC, RACE, HISPANIC, APOE, CLARiTI,
    `Amyloid PET (CL)`, `Tau PET (SUVR)`, `Hipp. atrophy (Z-score)`, Memory, `Exec Function`),
  type = all_continuous() ~ "continuous2",
  statistic = list(
    all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
    all_categorical() ~ "{n} ({p}%)"
  ),
  digits = all_continuous() ~ 1,
  missing_text = "(Missing)"
) %>%
  add_overall(last = TRUE) %>%
  bold_labels()
```

## Population-level severity trajectories

```{r}
#| label: severity-trajectory-prep
# update() returns the per-outcome hlme() sub-models with the joint
# estimates; predictY() on each gives class-specific mean trajectories
# (columns Ypred_class<k>, with lower./upper. bounds when draws = TRUE).
# update() re-evaluates the fit's stored call, which names the sub-models
# (m1 for K = 1, template_K<K>_Z<j> for K >= 2), so recreate them first.
m1 <- load_m1()
if (K_final > 1) invisible(make_templates(K_final, m1))
sub_models <- update(fit_final)

age_grid <- seq(min(wide_qz$age, na.rm = TRUE), max(wide_qz$age, na.rm = TRUE), length.out = 100)
ref_sex  <- names(sort(table(wide_qz$SEX), decreasing = TRUE))[1]
ref_educ <- mean(wide_qz$EDUC, na.rm = TRUE)

newdata_grid <- tibble(
  age_c = age_grid - age_center,
  SEX = factor(ref_sex, levels = levels(wide_qz$SEX)),
  EDUC = ref_educ) %>%
  add_age_basis()

severity_summary_time <- map_dfr(seq_along(sub_models), function(j) {
  pred <- predictY(sub_models[[j]], newdata = newdata_grid, var.time = "age_c",
    draws = TRUE)
  as_tibble(pred$pred) %>%
    mutate(age = age_grid) %>%
    pivot_longer(-age, names_to = "col", values_to = "value") %>%
    mutate(
      stat = case_when(
        startsWith(col, "lower.") ~ "lower",
        startsWith(col, "upper.") ~ "upper",
        TRUE ~ "estimate"),
      k = as.integer(sub(".*class([0-9]+)$", "\\1", col)),
      outcome = outcome_specs$outcome[j]) %>%
    select(-col) %>%
    pivot_wider(names_from = stat, values_from = value)
}) %>%
  left_join(outcome_map, by = "outcome") %>%
  mutate(
    latent_class = factor(k, levels = seq_len(K_final), labels = paste("Class", seq_len(K_final))),
    Outcome = recode(Outcome,
      "Amyloid PET (CL)" = "Amyloid PET", "Tau PET (SUVR)" = "Tau PET",
      "Hipp. atrophy (Z-score)" = "Hippocampal atrophy", "Exec Function" = "Executive function"),
    Outcome = factor(Outcome, levels = c("Amyloid PET", "Tau PET", "Hippocampal atrophy",
      "Memory", "Executive function"))
  )
```

```{r}
#| label: fig-severity-trajectory
#| fig-cap: "Modeled population-level severity trajectories by latent class over age (95% confidence bands), covariates at reference values (best-fitting model)."
ggplot(severity_summary_time, aes(x = age, y = estimate, colour = latent_class, fill = latent_class)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ Outcome, ncol = 3) +
  scale_colour_brewer(palette = "Set1", name = "Latent class") +
  scale_fill_brewer(palette = "Set1", name = "Latent class") +
  labs(
    title = "Modeled population-level severity trajectories",
    subtitle = "Fitted mean trajectory by class; covariates fixed at reference values",
    x = "Age", y = "Predicted severity (Z-scale)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Observed participant-level trajectories

```{r}
#| label: fig-spaghetti-plot
#| fig-cap: "Observed participant-level severity trajectories by latent class (best-fitting model)."
spaghetti_data <- ddl_qz %>%
  left_join(class_lookup %>% select(id, latent_class), by = "id", relationship = "many-to-one") %>%
  mutate(
    Outcome = recode(Outcome,
      "Amyloid PET (CL)" = "Amyloid PET", "Tau PET (SUVR)" = "Tau PET",
      "Hipp. atrophy (Z-score)" = "Hippocampal atrophy", "Exec Function" = "Executive function"),
    Outcome = factor(Outcome, levels = c("Amyloid PET", "Tau PET", "Hippocampal atrophy",
      "Memory", "Executive function"))
  )

ggplot(spaghetti_data, aes(x = age, y = Q, group = interaction(id, Outcome), colour = latent_class)) +
  geom_line(alpha = 0.40, linewidth = 0.45) +
  geom_point(alpha = 0.35, size = 0.6) +
  facet_wrap(~ Outcome, ncol = 3) +
  scale_colour_brewer(palette = "Set1", name = "Latent class") +
  scale_y_continuous(breaks = seq(0, 1, 0.25)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Observed participant-level severity trajectories",
    subtitle = "Observed severity quantile plotted against chronological age",
    x = "Age", y = "Observed severity quantile"
  ) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Model fit checks

Each sub-model in `sub_models` carries fitted values at the observed data
(`sub_models[[j]]$pred`), which can be compared with the observed $Z$
values, binned by age, per outcome and class, to check for systematic
under- or over-prediction.

# Limitations {#sec-limitations}

## The abandoned Bayesian latent-time joint model (Stan)

Before adopting the `multlcmm` approach used throughout this document, an
extended effort built two general-purpose Stan programs
(`ltjm_general.stan` for $K=1$, `lctjm_general.stan` for $K \ge 2$) around
a participant-specific latent time shift $\delta_i$, fit via CmdStanR (4
chains), and swept a grid over disease-time scale (chronological age vs.
time since first observation), fixed-effect intercept structure
(outcome-specific vs. common), class-specific intercept structure, and
outcome weighting. That approach converged cleanly at $K \le 2$ under
every configuration tried, but never reached acceptable convergence at
$K = 3$, across a long sequence of identification schemes for the
class-specific intercepts $\alpha_{jk}$, each motivated by the specific
failure of the one before it:

1. A single hard-zero intercept constraint on the same outcome
   (hippocampal atrophy) already ordering the class-defining slope. This
   converged at low $K$ but forced that outcome's fitted class means to
   coincide exactly at disease-time zero, producing visibly artificial
   trajectory intersections.
2. A sum-to-zero constraint across outcomes within each class
   ($\sum_j \alpha_{jk} = 0$), which avoided the artificial intersection
   but left every outcome's intercept completely unordered across classes.
   $K = 3$ failed to converge, with weak separation in the one ordered
   dimension (the slope anchor) letting the unordered intercepts swap
   between classes without technically violating any constraint.
3. Full ordering of every outcome's intercept (and, briefly, every
   outcome's slope too) across classes. This measurably regressed $K = 2$
   (previously clean; max_rhat 1.734, 130 parameters above $\hat R=1.01$
   under full slope ordering) and left $K=3$/$K=4$ comparably broken --
   zero divergent transitions alongside enormous Rhat/ESS, the textbook
   signature of chains trapped in different local optima
   (`diagnose_mode_trapping.R`) rather than a boundary or geometry problem.
4. A two-anchor hard-zero scheme: hippocampal atrophy's slope orders the
   classes as before, but a *different* outcome (memory) gets the
   hard-zero intercept constraint, avoiding the single-outcome overload of
   scheme 1. Combined with recentering the age axis at the 5th percentile
   of baseline age (rather than the sample mean) to relocate the resulting
   forced-trajectory-coincidence point from an arbitrary mid-course age to
   cohort entry -- a defensible assumption in growth-mixture modeling, not
   just a cosmetic fix, since $\delta_i$'s prior is centered at 0
   regardless of where the age axis itself is centered. This was the most
   scientifically defensible scheme tried and still did not converge at
   $K=3$: max_rhat 2.947, max_feff_rhat 2.947, 4285 parameters above
   $\hat R = 1.01$, 0 divergences.
5. Removing class-specific intercepts entirely
   ($\alpha_{jk} \equiv 0$ for every outcome and class), leaving only
   hippocampal atrophy's slope ordering to distinguish classes. This
   removes the whole identifiability sub-problem rather than proposing a
   sixth parameterization of it, and was the last step taken before this
   document moved to `multlcmm`; whether it resolves $K=3$ convergence on
   its own was not established before the switch.

Every step changed either the fitted convergence or the substantive claim
being made, which is why the mechanism, not just the outcome, is preserved
here rather than only the final negative result. `lctjm_general.stan`
still implements every scheme in the sequence (governed by data flags), so
any of them can be revisited without rewriting the Stan program, should a
reason arise to return to a latent-time joint-model framework. The
diagnostic scripts built to investigate this (`diagnose_mode_trapping.R`,
`explore_k3_modes.R`) remain useful for characterizing multi-modality in
any future MCMC-based model on this data.

## Limitations of the current (`mpjlcmm`) approach

- **Mostly cross-sectional PET.** About 85% of participants with amyloid or
  tau PET have a single scan, so class differences in the PET trajectories
  rest largely on between-person differences in level by age rather than
  within-person change. The PET sub-models are kept simple (random
  intercept, 2-df spline) for this reason.
- **Outcomes are linked only through the class.** Given the latent class,
  the outcome processes are independent: there is no correlation between a
  participant's random effects on different outcomes. A participant who
  is persistently high on amyloid is not modeled as more likely to be high
  on tau beyond what their class implies.
- **Local optima.** `gridsearch()`'s random starts reduce but do not
  eliminate the risk of a local optimum; a converged, BIC-competitive $K$
  that looks different under another seed or more starts deserves another
  look.
- **Fixed sub-model structure.** Spline df and random-effects structures
  are set per outcome in advance (`outcome_specs`) rather than compared.
- **BIC is not comparable with the WAIC/LOOIC values** of the abandoned
  Stan grid.

# Next steps: reintroducing latent time {#sec-next-steps}

The `mpjlcmm` model uses chronological age, so participants with the same
disease course but different onset ages follow different age curves, and
classes can partly reflect onset age rather than subtype. The abandoned Stan
models addressed this with a latent time shift $\delta_i$ but did not
converge at $K = 3$ (@sec-limitations). A converged `mpjlcmm` fit can make
that model easier to identify:

- **Initial values and priors for class trajectories:** the class-specific
  spline coefficients (`sub_models`) give starting values and informative
  priors for each class's trajectory shape per outcome.
- **Variance components:** the random-effect and residual variances per
  outcome suggest prior scales, and show how much between-person variation
  a time shift would need to absorb.
- **Class proportions and membership:** the class proportions and the
  APOE class-membership effects suggest priors for the mixture weights;
  posterior class probabilities (`class_lookup`) can initialize class
  assignments, or the latent-time model can be fitted within each class.
- **Parameterization:** which outcomes separate the classes (where the class
  trajectories differ most) suggests which outcome should anchor the class
  ordering or the time scale.

# Session information

```{r}
#| label: session-info
sessionInfo()
```
