---
title: "Latent-Class, Latent-Time Joint Models for CLARiTI: Candidate Model Comparison"
subtitle: "Model space, fitting workflow, WAIC/LOOIC comparison, and summary of the best-fitting model"
author: "Li, Donohue, Langford, Insel, et al."
date: today
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

<!--
This QMD reorganizes the exploratory work in `lctjm-clariti-report.Rmd` into a
single pipeline that fits, screens, and compares a pre-specified grid of
latent-time / latent-class joint models, then summarizes the best-fitting
model. It is written to be run end-to-end by the user; no model fitting is
executed as part of drafting this file (fitting chunks are `eval: false` and
read from cached `.RDS` files once you have run them).

Two new Stan files generalize the four hand-written models in the original
draft (`ltjm.stan`, `lctjm.stan`, `lctjm-noncentered-delta.stan`,
`lctjm-noncentered-delta-class-specific.stan`) into two files with
data-driven toggles, so the whole candidate grid below is fit from just two
compiled Stan programs:

  - `ltjm_general.stan`   -- K = 1 (single latent class / no mixture)
  - `lctjm_general.stan`  -- K >= 2 (latent-class mixture)

Both accept an `outcome_specific_intercept` flag and a `weight` vector;
`lctjm_general.stan` additionally accepts a `class_specific_intercept` flag.
See the Methods section for the full model definitions.
-->

```{r}
#| label: packages
# cmdstanr/posterior/bayesplot/loo/tidybayes/modelr were needed for the
# abandoned Bayesian Stan grid (@sec-limitations) and are no longer used;
# lcmm and splines are the new fitting/spline-basis dependencies.
library(lcmm)
library(splines)
library(tidyverse)
library(patchwork)
library(knitr)
library(kableExtra)
library(data.table)
library(gtsummary)

set.seed(20260910)

options(htmltools.dir.version = FALSE, knitr.kable.NA = "")

knitr::opts_chunk$set(
  echo = FALSE,
  tidy = FALSE,
  collapse = TRUE,
  fig.path = "",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  out.extra = "",
  out.width = "\\linewidth",
  crop = TRUE,
  fig.pos = "htbp",
  cache = FALSE,
  comment = "",
  fig.dpi = 300
)

dir.create("fits", showWarnings = FALSE, recursive = TRUE)
```

```{r}
#| label: ggplot-setup
theme_jama <- function(base_size = 10) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      plot.title = element_text(size = base_size + 1, face = "bold", hjust = 0, margin = margin(b = 10)),
      plot.subtitle = element_text(size = base_size, hjust = 0, margin = margin(b = 10)),
      plot.caption = element_text(size = base_size - 1, hjust = 1, margin = margin(t = 10)),
      axis.title = element_text(size = base_size, face = "bold"),
      axis.text = element_text(size = base_size - 1, color = "black"),
      axis.line = element_line(color = "black", linewidth = 0.5),
      axis.ticks = element_line(color = "black", linewidth = 0.5),
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.title = element_text(size = base_size, face = "bold"),
      legend.text = element_text(size = base_size - 1),
      legend.key = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.position = "bottom",
      strip.background = element_blank(),
      strip.text = element_text(size = base_size - 2, face = "bold", vjust = 1, color = "black"),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

jama_colors <- c("#374E55", "#DF8F44", "#00A1D5", "#B24745", "#79AF97", "#6A6599", "#80796B")
scale_color_jama <- function(...) scale_color_manual(values = jama_colors, ...)
scale_fill_jama  <- function(...) scale_fill_manual(values = jama_colors, ...)
```

```{r}
#| label: ecdf-functions
ecdf.func <- function(x, weights = NULL,
  type = c("i/n", "(i-1)/(n-1)", "i/(n+1)"),
  normwt = TRUE, na.rm = TRUE,
  method = "linear", rule = 2, ...) {
  type <- match.arg(type)
  ecdf.data <- Hmisc::wtd.Ecdf(x, weights = weights, type = type,
    normwt = normwt, na.rm = na.rm)
  with(ecdf.data, approxfun(x, ecdf, method = method, na.rm = na.rm, rule = rule, ties = mean, ...))
}
```

# Introduction

An earlier exploratory analysis (`lctjm-clariti-report.Rmd`) fit a sequence
of hand-tuned latent-time and latent-class joint models to five harmonized
CLARiTI outcomes (hippocampal atrophy, amyloid PET, tau PET, memory, and
executive function) and found evidence, via LOOIC, for three latent classes
with class-specific intercepts. A systematic follow-up built two
general-purpose Stan programs (`ltjm_general.stan`/`lctjm_general.stan`)
around a shared participant-specific *latent time shift* $\delta_i$ and
swept a candidate grid over disease-time scale, fixed-effect intercept
structure, class-specific-intercept structure, and outcome weighting. That
line of work is preserved in full in @sec-limitations, including the exact
convergence diagnostics from each identifiability scheme tried: every
variant converged at $K \le 2$ but none reached acceptable convergence at
$K = 3$ (best case: max_rhat 2.947, 4285 parameters above $\hat R = 1.01$),
even after removing class-specific intercepts entirely and leaving only
hippocampal atrophy's slope to distinguish classes.

This document now takes a different approach: a **latent-class mixed
model, jointly across the same five outcomes, with no latent time shift**.
Each participant's own chronological age at each observation is used
directly as the time variable (no $\delta_i$, no disease-time
recentering), and each outcome's trajectory over age is modeled with a
natural cubic spline (3 degrees of freedom) rather than a single linear
slope, so curvature in disease progression doesn't have to be captured
through $\delta_i$ and a linear rate the way the joint-model approach
required. This is fit with `lcmm::multlcmm()` (Proust-Lima et al.), a
maximum-likelihood latent-class mixed model for multiple longitudinal
outcomes, rather than with Stan/CmdStanR -- see @sec-methods for why
`multlcmm` (not `lcmm::Jointlcmm`, which mandatorily couples the
longitudinal model to a time-to-event submodel this analysis has no use
for, and not `rstanarm::stan_jm`, which has the same mandatory
survival-linkage and, more fundamentally, no discrete latent-class
mechanism at all) is the right tool for this specification. If a $K$
converges cleanly here, the natural next step -- not yet implemented in
this document -- is refitting `ltjm_general.stan` (the $K=1$ latent-time
model, which converges without difficulty) separately within each
resulting class, now that the latent-class structure would already be
resolved by `multlcmm` rather than needing to be estimated jointly with
$\delta_i$ inside the same Stan model.

This document lays out the `multlcmm` model specification, provides the
data-assembly and model-fitting code, sweeps the number of latent classes
($K = 1, \dots, 4$), compares candidates using BIC and posterior
classification quality (mean assignment probability, entropy), and
summarizes the best-fitting model. A **Methods** section (@sec-methods)
gives the full statistical description suitable for a manuscript.

Fitting is far cheaper than the abandoned MCMC grid (maximum-likelihood
optimization rather than sampling; @sec-fitting), but multiple random
starting values per $K$ are still needed to guard against local optima
(`lcmm::gridsearch()`), so the fitting chunks below are still marked
`eval: false`. Run them once in an interactive session, then re-render for
the comparison and summary sections, which read the cached `.RDS` fit
objects from the `fits/` directory.

# Data {#sec-data}

## Long-format data assembly

The long-format participant-visit dataset (demographics, hippocampal
atrophy, amyloid PET, tau PET, and harmonized cognitive composites) is
built from the source NACC/CLARiTI tables exactly as in the original
exploratory report. That derivation is reproduced here for provenance but
is not re-run (`eval: false`); it produced
`clariti-long-atn-union-data.RDS`, i.e. `claritilongatnuniondata.RDS`, which
is the entry point for this document.

```{r}
#| label: clariti-long-data-setup
#| eval: false

library(NACCADRC)
data(uds_mri, mrisbm, taupetnpdka, amyloidpetgaain, phc_cognition, uds_ftldlbd)

# demographics ----

demog <- uds_ftldlbd %>%
  mutate(BIRTHDATE = as.IDate(paste(BIRTHYR, BIRTHMO, 15, sep = '-'))) %>%
  dplyr::arrange(NACCID, NACCVNUM) %>%
  select(NACCID, Dx = NACCUDSD, BIRTHDATE, RACE, SEX = NACCSEX, EDUC,
    HISPANIC = NACCHISP, APOE = NACCAPOE) %>%
  mutate(
    APOE = case_when(
      APOE == 1 ~ 'e3e3',
      APOE == 2 ~ 'e3e4',
      APOE == 3 ~ 'e3e2',
      APOE == 4 ~ 'e4e4',
      APOE == 5 ~ 'e4e2',
      APOE == 6 ~ 'e2e2',
      APOE == 9 ~ NA) %>% 
      factor(levels = c('e2e2', 'e3e2', 'e4e2', 'e3e3', 'e3e4', 'e4e4'))) %>%
  dplyr::group_by(NACCID) %>%
  tidyr::fill(.direction = 'downup') %>%
  filter(!is.na(BIRTHDATE), !is.na(APOE), APOE != 'Missing/ unknown/ not assessed') %>%
  dplyr::distinct(NACCID, .keep_all = TRUE) %>%
  mutate(across(where(is.factor), fct_drop)) %>%
  dplyr::ungroup()

# hippocampal volume ----
mri_hipp <- demog %>%
  dplyr::left_join(
    mrisbm %>%
      select(PROJECT, NACCID, DATE = SCANDT, HIPPOCAMPUS, ICV = CEREBRUMTCV) %>%
      dplyr::bind_rows(
        uds_mri %>%
          select(NACCID, MRIYR, MRIMO, MRIDY, HIPPOCAMPUS = HIPPOVOL, ICV = NACCICV) %>%
          filter(HIPPOCAMPUS != 88.8888, ICV != 9999.999, !is.na(HIPPOCAMPUS)) %>%
          mutate(
            DATE = as.IDate(paste(MRIYR, MRIMO, MRIDY, sep = '-')),
            PROJECT = 'Mixed protocol')),
    by = 'NACCID') %>%
  dplyr::distinct(NACCID, DATE, .keep_all = TRUE) %>%
  filter(!is.na(HIPPOCAMPUS), !is.na(ICV))

mri_icv_mod <- lm(HIPPOCAMPUS ~ scale(ICV), data = mri_hipp)
mri_hipp$Hipp_res <- mri_icv_mod$coefficients[['(Intercept)']] + mri_icv_mod$residuals
mri_hipp$Hipp_atrophy <- scale(-mri_hipp$Hipp_res)[, 1]

# amyloid PET ----
amy_pet <- demog %>%
  dplyr::left_join(
    amyloidpetgaain %>%
      dplyr::arrange(NACCID, SCANDATE, desc(PROCESSDATE)) %>%
      dplyr::distinct(NACCID, SCANDATE, .keep_all = TRUE) %>%
      select(NACCID, DATE = SCANDATE, AMYLOID_STATUS, CENTILOIDS,
        Amyloid_TRACER = TRACER, Amyloid_PROJECT = PROJECT) %>%
      dplyr::distinct(NACCID, DATE, .keep_all = TRUE),
    by = 'NACCID')

# tau PET ----
 tau_pet <- demog %>%
  dplyr::left_join(
    taupetnpdka %>%
      dplyr::arrange(NACCID, SCANDATE, desc(PROCESSDATE)) %>%
      dplyr::distinct(NACCID, SCANDATE, .keep_all = TRUE) %>%
      select(NACCID, DATE = SCANDATE, Tau_PET = META_TEMPORAL_SUVR,
        Tau_TRACER = TRACER, Tau_PROJECT = PROJECT) %>%
      dplyr::distinct(NACCID, DATE, .keep_all = TRUE),
    by = 'NACCID') %>%
  filter(!is.na(Tau_PET)) %>%
  mutate(age = as.numeric(as.Date(DATE) - as.Date(BIRTHDATE)) / 365.25) %>% mutate(SEX = droplevels(factor(SEX)),
    Dx  = droplevels(factor(Dx)))

tau_PET_fit <- lme(Tau_PET ~ Tau_TRACER + SEX + age + Dx, data = tau_pet,
  random = ~ 1 | NACCID,
  weights = varIdent(form = ~ 1 | Tau_TRACER))

set.seed(42)

tau_pet$Tau_PET_jitter <- tau_pet$Tau_PET + rnorm(nrow(tau_pet), 0, 1e-6)

tau_pet$Tau_PET_ComBat <- as.numeric(
  sva::ComBat(
    dat = rbind(Tau_PET = tau_pet$Tau_PET,
                Tau_PET_jitter = tau_pet$Tau_PET_jitter),
    batch = factor(tau_pet$Tau_TRACER),
    mod = model.matrix(~ SEX + age + Dx, data = tau_pet)
  )["Tau_PET", ])

# cognition ----
cog <- demog %>%
  left_join(
    phc_cognition %>%
      select(NACCID, NACCVNUM, PHC_MEM, PHC_EXF, PHC_LAN) %>%
      left_join(
        uds_ftldlbd %>% select(NACCID, NACCVNUM, DATE = VISITDATE),
        by = c('NACCID', 'NACCVNUM')
      ),
    by = 'NACCID'
  ) %>%
  pivot_longer(PHC_MEM:PHC_LAN, names_to = 'Outcome', values_to = 'Y') %>%
  mutate(
    Y = -Y,
    outcome = case_when(
      Outcome == 'PHC_MEM' ~ 4,
      Outcome == 'PHC_EXF' ~ 5,
      Outcome == 'PHC_LAN' ~ 6,
      TRUE ~ NA_real_
    ),
    Outcome = case_when(
      Outcome == 'PHC_MEM' ~ 'Memory',
      Outcome == 'PHC_EXF' ~ 'Exec Function',
      Outcome == 'PHC_LAN' ~ 'Language',
      TRUE ~ Outcome
    )
  )

# long data ----

ddl <- mri_hipp %>%
  select(NACCID, Dx, BIRTHDATE, RACE, SEX, EDUC, HISPANIC, APOE, DATE, Y = Hipp_atrophy) %>%
  mutate(outcome = 1, Outcome = 'Hipp. atrophy (Z-score)') %>%
  bind_rows(
    amy_pet %>%
      select(NACCID, Dx, BIRTHDATE, RACE, SEX, EDUC, HISPANIC, APOE, DATE, Y = CENTILOIDS) %>%
      mutate(outcome = 2, Outcome = 'Amyloid PET (CL)')) %>%
  bind_rows(
    tau_pet %>%
      select(NACCID, Dx, BIRTHDATE, RACE, SEX, EDUC, HISPANIC, APOE, DATE, Y = Tau_PET_ComBat) %>%
      mutate(outcome = 3, Outcome = 'Tau PET (SUVR)')) %>%
  bind_rows(
    cog %>% select(NACCID, Dx, BIRTHDATE, RACE, SEX, EDUC, HISPANIC, APOE, DATE, 
      Y, outcome, Outcome)) %>%
  filter(!is.na(Y), !is.na(DATE)) %>%
  group_by(NACCID) %>%
  mutate(
    age = as.numeric(DATE - BIRTHDATE) / 365.25,
    age0 = min(age, na.rm = TRUE),
    years = age - age0) %>%
  ungroup()

complete_covariates <- ddl %>%
  arrange(NACCID, years) %>%
  select(NACCID, age0, SEX, APOE, EDUC) %>%
  filter(SEX %in% c('Female', 'Male')) %>%
  distinct() %>%
  na.omit()

ddl <- ddl %>%
  filter(NACCID %in% complete_covariates$NACCID) %>%
  mutate(
    id = factor(NACCID) %>% as.numeric(),
    across(where(is.factor), fct_drop)
  ) %>%
  arrange(id, outcome) %>%
  group_by(id) %>%
  mutate(years_c = years ) %>% #consider time centering with years_c = years - mean(years)
  ungroup()

## Save long data ----

saveRDS(ddl, 'clariti-long-data.RDS')

## Save long atn complete data ----

id.a <- filter(ddl, Outcome == 'Amyloid PET (CL)') %>% pull(NACCID)
id.t <- filter(ddl, Outcome == 'Tau PET (SUVR)') %>% pull(NACCID)
id.n <- filter(ddl, Outcome == 'Hipp. atrophy (Z-score)') %>% pull(NACCID)
id.c <- filter(ddl, Outcome == 'Memory') %>% pull(NACCID)

id.int <- intersect(id.a, id.t) %>% intersect(id.n) %>% intersect(id.c) %>% unique()

ddl_atn_complete <- ddl %>%
  filter(NACCID %in% id.int) %>%
  mutate(id = factor(NACCID) %>% as.numeric())

saveRDS(ddl_atn_complete, 'clariti-long-atn-complete-data.RDS')

## Save long atn union data ----

id.union <- c(id.a, id.t, id.n) %>% unique() %>% intersect(id.c)

ddl_atn_union <- ddl %>%
  filter(NACCID %in% id.union) %>%
  mutate(id = factor(NACCID) %>% as.numeric())

saveRDS(ddl_atn_union, 'clariti-long-atn-union-data.RDS')
```

```{r}
#| label: load-data
ddl_atn_union <- readRDS('clariti-long-atn-union-data.RDS')
ddl <- ddl_atn_union
```

## Standardizing outcomes to a common severity scale

Because the five outcomes are on incommensurable scales (Z-scored
hippocampal atrophy, amyloid centiloids, tau SUVR, and harmonized cognitive
composites), each is mapped to a common severity metric before joint
modeling: a diagnosis-reweighted empirical CDF converts each raw outcome to
a severity quantile $Q \in (0,1)$ (reweighted so that Cognitively
Normal/MCI/Dementia are represented as in the general population rather
than in this enriched sample), and $Z = \Phi^{-1}(Q)$ places all outcomes
on a common standard-normal severity scale. This transform, and the
outcome coding, are unchanged from the original draft.

```{r}
#| label: clariti-q-z-scores
ddl_ecdf <- ddl %>%
  mutate(
    Dx_ecdf = forcats::fct_collapse(Dx, MCI = c("MCI", "Impaired-not-MCI"))
  )

ecdf_maps_collapsed <- ddl_ecdf %>%
  dplyr::count(Outcome, Dx_ecdf, name = "n_dx") %>%
  group_by(Outcome) %>%
  mutate(prop_dx = n_dx / sum(n_dx)) %>%
  ungroup() %>%
  right_join(ddl_ecdf, by = c("Outcome", "Dx_ecdf")) %>%
  mutate(weight = 1 / prop_dx) %>%
  group_by(Outcome) %>%
  summarise(
    Y_to_Q = list(ecdf.func(x = Y, weights = weight, type = "i/(n+1)")),
    .groups = "drop"
  )

ddl_qz <- ddl_ecdf %>%
  left_join(ecdf_maps_collapsed, by = "Outcome") %>%
  mutate(
    Q = map2_dbl(Y_to_Q, Y, ~ .x(.y)),
    Z = qnorm(Q)
  ) %>%
  arrange(id, outcome)
```

## Outcome set and coding

As in the original draft, Language is dropped for data-completeness
reasons, leaving $J = 5$ outcomes. Unlike the abandoned Stan approach
(@sec-limitations), no outcome needs to be designated an "anchor": there is
no hard identifying constraint to place anywhere. `multlcmm` handles
class-label ambiguity differently from the Bayesian MCMC setting -- there
is no cross-chain label-switching to prevent, since each `multlcmm` fit is
a single maximum-likelihood optimization, not a collection of chains that
have to agree on which class is "class 1." The corresponding risk in the
ML setting is convergence to a local rather than global optimum, which
`lcmm::gridsearch()` addresses by trying many random starting values and
keeping the best log-likelihood (@sec-fitting) rather than by constraining
any one parameter's sign or ordering.

```{r}
#| label: outcome-coding
recode_outcome <- function(dat) {
  dat %>%
    filter(Outcome != "Language") %>%
    mutate(
      outcome = case_when(
        Outcome == "Hipp. atrophy (Z-score)" ~ 1,
        Outcome == "Amyloid PET (CL)"        ~ 2,
        Outcome == "Tau PET (SUVR)"          ~ 3,
        Outcome == "Memory"                  ~ 4,
        Outcome == "Exec Function"           ~ 5,
        TRUE ~ NA_real_
      )
    ) %>%
    arrange(id, outcome, years)
}

ddl <- recode_outcome(ddl)
ddl_qz <- recode_outcome(ddl_qz)

J <- length(unique(ddl_qz$outcome))

outcome_map <- ddl_qz %>% distinct(outcome, Outcome) %>% arrange(outcome)
```

# Candidate model space {#sec-grid}

## Comparison axis

| Axis | Levels | `multlcmm` mechanism |
|---|---|---|
| Number of latent classes ($K$) | 1, 2, 3, 4 | `ng` argument |

Age-at-observation is used directly as the time variable for every $K$
(no disease-time-scale axis: there is no $\delta_i$ left to recenter
against, so the age-recentering identifiability concern that drove the
Stan approach's "Disease-time scale" axis, @sec-limitations, doesn't arise
here). The trajectory shape over age is a natural cubic spline with 3
degrees of freedom, fixed rather than swept (@sec-methods) -- 2 df was
also considered; 3 was chosen for enough flexibility to capture
non-monotone or accelerating decline without adding enough parameters per
class to reintroduce the kind of convergence fragility the Stan grid ran
into. Every outcome gets its own intercept and its own spline coefficients
(via `contrast()`, @sec-methods) rather than sharing one common-intercept
vs. outcome-specific-intercept toggle, so that axis from the Stan grid is
also gone -- it's not a choice being fixed one way, it's a modeling
question that no longer has the same two discrete answers. There is
likewise no outcome-weighting axis: `multlcmm` has no case-weighting
argument, and unlike the Stan grid's pseudo-likelihood weighting, there's
no straightforward way to add one without changing the likelihood in ways
that would need to be implemented and validated from scratch. Only $K$
remains a free axis for comparison below.

## Specification grid

```{r}
#| label: spec-grid
# Only K is being compared; see the "Comparison axis" note above for why
# every axis from the abandoned Stan grid (disease-time scale, intercept
# structure, class-specific intercepts, weighting) either no longer
# applies or no longer has a discrete on/off form under multlcmm.
spec_grid <- tibble::tibble(K = 1:4) %>%
  mutate(model_id = paste0("multlcmm_K", K))

nrow(spec_grid)
```

```{r}
#| label: tab-spec-grid
kable(
  spec_grid,
  caption = "Candidate specification grid: number of latent classes only."
)
```

# Statistical methods {#sec-methods}

## Model {#sec-methods-model}

For participant $i$, outcome $j = 1, \dots, J$, and observation $v$ at
chronological age $a_{iv}$, `multlcmm` models the (already
severity-standardized, @sec-data) outcome as

$$
Y_{ijv} = \beta_{0j} + f_{jk(i)}(a_{iv}) + \mathbf{x}_i^{\mathsf T}\boldsymbol\beta_j
  + u_{i} + v_i\, a_{iv} + w_{ij} + \varepsilon_{ijv}, \qquad
  \varepsilon_{ijv} \sim \operatorname{Normal}(0, \sigma_j^2),
$$

where $k(i) \in \{1, \dots, K\}$ is participant $i$'s (unobserved) latent
class, $\beta_{0j}$ is outcome $j$'s fixed intercept, $\mathbf x_i$ contains
sex and education, and $f_{jk}(a) = \sum_{m=1}^{3} \gamma_{jkm}\, N_m(a)$ is
a natural-cubic-spline (3 df, basis functions $N_m$) trajectory over age,
specific to *both* outcome $j$ and class $k$. Two random effects are
shared across all $J$ outcomes for a given participant -- a random
intercept $u_i$ and a random linear slope on age $v_i$, both
$\operatorname{Normal}(0, \cdot)$ with a freely estimated covariance,
representing the "common latent process" `multlcmm` is built around -- and
$w_{ij} \sim \operatorname{Normal}(0, \cdot)$ is an *outcome-specific*
random intercept (`randomY = TRUE`), letting participant $i$ run
persistently high or low on outcome $j$ specifically, beyond whatever the
shared process and the class-specific spline already explain.

**Why outcome-specific splines, not one shared latent trajectory.**
`multlcmm`'s default and most common use (e.g. modeling several cognitive
test scores as noisy indicators of one underlying cognitive ability) treats
the outcomes as different-scaled measurements of a *single* latent
trajectory, with only their loadings/link functions allowed to differ. That
default would be a poor fit for this application: amyloid, tau, hippocampal
atrophy, and cognition are understood to become abnormal at different,
overlapping stages of disease (the premise behind biomarker-staging /
event-based-disease-progression models), not as proportional readouts of
one common severity dimension. Wrapping the spline term in `mixture()`
*and* `contrast()` (rather than leaving it in `fixed`/`mixture` alone) gives
every outcome $j$ its own class-specific spline coefficients
$\gamma_{jkm}$, so each outcome's trajectory shape -- including whether and
when it departs from flat -- is estimated separately, and classes are free
to differ from each other differently on different outcomes. This is the
`multlcmm`-native way of reproducing what the abandoned Stan model did by
giving every outcome its own `beta_s[j,k]` slope (@sec-limitations); the
shared random intercept/slope ($u_i, v_i$) is what still ties the $J$
outcomes together for a given participant, rather than $\delta_i$.

## Class membership model

Class membership is modeled with a multinomial logistic regression on APOE
genotype (`classmb = ~ APOE`), the same predictor used throughout the
abandoned Stan grid, with class 1 as the reference category in `lcmm`'s
parameterization (confirm against a given fit's `summary()`/`postprob()`
output, since which raw class index ends up "class 1" depends on
`gridsearch()`'s best run, not on any constraint imposed here).

## Local optima and multiple starting values {#sec-methods-optima}

A maximum-likelihood latent-class mixture does not have the Stan
grid's specific failure mode (chains disagreeing about where the posterior
mass is; @sec-limitations) because there's no posterior and no chains --
each fit is one optimization run. It has a different, related risk
instead: the likelihood surface for a $K$-class mixture is generally
multimodal, so a single optimization run from one starting point can
converge to a local rather than global maximum, and different starting
points can converge to genuinely different class solutions (analogous in
spirit to the Stan grid's mode-trapping, but a property of the likelihood
surface itself rather than of chain initialization). `lcmm::gridsearch()`
addresses this directly: it fits the model from many random perturbations
of an initial (typically $K=1$ or $K-1$) solution and keeps the run with
the highest converged log-likelihood, playing the same role dispersed
chain inits played for the Stan grid, but resolved by taking the best
optimum found rather than by checking whether independent runs agree.
@sec-fitting uses it for every $K \ge 2$ fit below.

## Estimation

`multlcmm` estimates parameters by maximum likelihood via a
Marquardt-Levenberg algorithm (a modified Newton-Raphson with numerical
first and second derivatives), not MCMC -- there are no priors to specify
and no posterior draws to summarize; parameter uncertainty is instead
reported via the asymptotic (Wald) standard errors from the converged
Hessian, and `predictY(..., draws = TRUE)` approximates confidence bands
for fitted trajectories via Monte Carlo draws from that same asymptotic
parameter distribution. Convergence is indicated by `multlcmm`'s own
convergence code (1 = converged) together with a positive-definite
Hessian; @sec-fitting screens on both, and additionally refits every
$K \ge 2$ specification from `gridsearch()`'s multiple starts
(@sec-methods-optima) rather than accepting a single run's convergence
code alone.

## Model comparison

Candidate $K$ are compared using BIC ($-2\times\text{log-likelihood} +
p\log N$, $p$ = number of parameters, $N$ = number of participants; lower
is better) rather than WAIC/LOOIC, since there is no posterior to compute
either from. BIC is reported alongside two classification-quality
diagnostics that a fit statistic alone can miss: mean posterior
class-assignment probability (averaged over participants, and within each
class) and the normalized entropy of the posterior class-membership
probabilities (`lcmm::postprob()`), both of which indicate how cleanly
participants are actually being separated into classes, as opposed to how
well the model fits the data overall -- a low-BIC model whose class
probabilities are all close to $1/K$ for many participants has found $K$
components that improve fit without meaningfully classifying anyone, which
BIC alone would not flag.

# Model fitting {#sec-fitting}

## Compute and storage strategy {#sec-fitting-strategy}

Maximum-likelihood optimization is far cheaper than the abandoned MCMC
grid -- a single `multlcmm` run takes seconds to low minutes rather than
the tens of minutes to hours a 4-chain CmdStanR fit needed -- but
`gridsearch()`'s whole point is to run that optimization from many (here,
50) random starts per $K \ge 2$ candidate, so total compute is still
non-trivial and worth caching. `fit_or_load_multlcmm()` below caches by
`model_id` to a single `.RDS` file per $K$ under `fits/`; unlike the
two-tier full-vs-scorecard split the abandoned Stan grid needed
(@sec-limitations), a converged `multlcmm`/`gridsearch` object has no
posterior draws array to strip out -- it's already just parameter
estimates, the Hessian, and fitted values -- so one cache file per model is
enough.

## Helper functions

```{r}
#| label: helper-multlcmm-data
# Reshape ddl_qz from long-by-outcome (one row per (id, outcome,
# observation), as the abandoned Stan grid needed) into the wide format
# multlcmm expects: one row per (id, observation time), with each
# outcome's severity Z-score in its own column (NA where that outcome
# wasn't measured at that particular age). Outcomes measured at visits
# that don't line up exactly in age across outcomes (plausible here, since
# PET imaging and cognitive testing aren't always done on the same day)
# simply produce more, sparser rows rather than failing -- multlcmm
# handles a fully unbalanced multi-outcome panel like this natively; there
# is no requirement that every outcome be measured at every row's age.
# values_fn guards against two observations of the same outcome landing on
# the exact same age for the same id (a rounding collision); this should
# be rare, but check `nrow(wide) == nrow(distinct(wide))` after building it
# for a given dataset to confirm it isn't silently averaging anything real.
build_multlcmm_data <- function(ddl_qz) {
  ddl_qz %>%
    transmute(id, outcome, age, SEX, APOE, EDUC, Z,
      outcome_lab = paste0("Z", outcome)) %>%
    tidyr::pivot_wider(
      id_cols = c(id, age, SEX, APOE, EDUC),
      names_from = outcome_lab,
      values_from = Z,
      values_fn = ~ mean(.x, na.rm = TRUE)
    ) %>%
    arrange(id, age)
}

wide_qz <- build_multlcmm_data(ddl_qz)
stopifnot(nrow(wide_qz) == nrow(distinct(wide_qz, id, age)))

# age_c: age centered at the sample mean, used everywhere in the fitting
# formulas below IN PLACE OF raw age. This is purely a numerical-
# conditioning fix, not a re-introduction of the abandoned Stan model's
# disease-time recentering (that was about relocating a hard-zero
# identifiability constraint; there is no such constraint here). A random
# intercept + random slope on raw chronological age (typically ~60-90 in
# this cohort) strongly correlates the two random effects and is a classic
# cause of Marquardt-Levenberg convergence failure; centering removes that
# without changing the model. age_center is kept so any newdata passed to
# predictY() later can be centered the same way, and so age_c can always
# be converted back to real age (age = age_c + age_center) for plotting.
age_center <- mean(wide_qz$age, na.rm = TRUE)
wide_qz <- wide_qz %>% mutate(age_c = age - age_center)
```

```{r}
#| label: outcome-coverage-summary
#| tbl-cap: "Per-outcome data coverage: how much information each outcome's contrast()-ed spline actually has to work with."
# Added while diagnosing multlcmm non-convergence (@sec-fitting): giving
# every outcome its own fully contrast()-ed age spline (@sec-methods-model)
# assumes there's enough data, spread widely enough over age, within EACH
# outcome to estimate that outcome's own 3 spline coefficients (x K classes
# for the mixture term). If one or two outcomes (plausibly amyloid/tau PET,
# which are typically collected far less often than cognitive testing or
# structural MRI) are measured on a small fraction of subjects or over a
# narrow age range, their contrasted spline coefficients can be weakly
# identified -- a likely driver of the Hessian/non-convergence problems,
# distinct from (though related to) missingness itself. Check this BEFORE
# assuming the fix is architectural (dropping contrast() for sparse
# outcomes, reducing df, or outcome-specific exclusion) rather than just
# optimizer tuning.
outcome_coverage <- ddl_qz %>%
  filter(!is.na(Z)) %>%
  group_by(outcome, Outcome) %>%
  summarise(
    n_subjects = n_distinct(id),
    n_obs = n(),
    age_min = min(age, na.rm = TRUE),
    age_max = max(age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(pct_of_subjects = round(100 * n_subjects / n_distinct(ddl_qz$id), 1)) %>%
  arrange(desc(n_subjects))

kable(outcome_coverage, digits = 1)
```

```{r}
#| label: helper-multlcmm-fit
# multlcmm's own formula interface, not a hand-built stan_data list, is
# now the API surface that matters -- see @sec-methods-model for why the
# age-trajectory spline is wrapped in BOTH mixture() (class-varying) and
# contrast() (outcome-varying), rather than left as a term shared across
# classes and/or outcomes the way multlcmm's more common single-latent-
# process use case would leave it. Double-check this expands to the
# intended per-outcome, per-class spline coefficients via
# summary(fit)/fit$Xnames before trusting it on a first run -- this is a
# less common way to use multlcmm than the package's own vignette
# examples, and it hasn't been run here (the standing constraint on this
# whole project is that Claude never fits these models itself).
outcome_names_z <- paste0("Z", sort(unique(ddl_qz$outcome)))
fixed_lhs <- paste(outcome_names_z, collapse = " + ")
# age_c (age centered at the sample mean; see helper-multlcmm-data), not
# raw age, is what actually enters every formula below -- purely for
# optimizer conditioning, not a change to the model.
fixed_formula <- as.formula(paste(
  fixed_lhs, "~ contrast(splines::ns(age_c, df = 3)) + SEX + EDUC"
))
mixture_formula <- ~ contrast(splines::ns(age_c, df = 3))
classmb_formula <- ~ APOE

# Fails loudly and immediately on a non-converged fit rather than caching
# it -- letting a bad K = 1 fit sit in the cache silently was exactly what
# turned into a confusing gridsearch() error two steps downstream at K = 2
# ("The model minit did not converge"), which is gridsearch() refusing to
# perturb a minit that never converged in the first place. conv codes (see
# ?multlcmm): 1 = converged; 2 = max iterations reached without meeting the
# convergence criteria; 3 = a convergence criterion was met but the
# Hessian isn't positive definite; 4/5 = other numerical problems.
check_converged <- function(fit, model_id) {
  if (fit$conv != 1) {
    stop(sprintf(paste(
      "%s did not converge (conv = %d; see ?multlcmm for what this code",
      "means). Not caching this result. Things worth trying, roughly in",
      "order of how often they help: (1) confirm age_c (centered), not raw",
      "age, is what's actually in the formula -- an uncentered time",
      "variable in a random-slope model is a common cause of exactly this;",
      "(2) raise maxiter further; (3) temporarily simplify the",
      "random-effects structure (random = ~1 only, or randomY = FALSE) to",
      "confirm a simpler model converges, then reintroduce complexity one",
      "piece at a time to isolate which piece is causing it -- the same",
      "strategy used throughout the abandoned Stan grid's identifiability",
      "search (@sec-limitations)."
    ), model_id, fit$conv))
  }
  fit
}

fit_or_load_multlcmm <- function(model_id, K, data, n_starts = 50, seed = 20260101) {
  path <- file.path("fits", paste0(model_id, ".RDS"))
  if (file.exists(path)) return(readRDS(path))

  set.seed(seed)

  if (K == 1) {
    fit <- multlcmm(
      fixed = fixed_formula,
      random = ~ 1 + age_c,
      randomY = TRUE,
      subject = "id",
      link = rep("linear", J),
      maxiter = 150,
      data = data
    )
    check_converged(fit, model_id)
  } else {
    m1_path <- file.path("fits", "multlcmm_K1.RDS")
    stopifnot(
      "Fit and cache K = 1 first; K >= 2 gridsearch() starts are dispersed
      perturbations of it (minit), not a fresh random start." =
        file.exists(m1_path)
    )
    m1 <- readRDS(m1_path)

    fit <- gridsearch(
      rep = n_starts, maxiter = 30, minit = m1,
      multlcmm(
        fixed = fixed_formula,
        mixture = mixture_formula,
        classmb = classmb_formula,
        random = ~ 1 + age_c,
        randomY = TRUE,
        ng = K,
        subject = "id",
        link = rep("linear", J),
        data = data
      )
    )
    check_converged(fit, model_id)
  }

  dir.create("fits", showWarnings = FALSE)
  saveRDS(fit, path)
  fit
}
```

## Stage 1: screen $K$ {#sec-stage1}

```{r}
#| label: stage1-fit
stage1_fits <- purrr::map(1:4, function(k) {
  fit_or_load_multlcmm(model_id = paste0("multlcmm_K", k), K = k, data = wide_qz)
})
names(stage1_fits) <- paste0("K", 1:4)
```

```{r}
#| label: stage1-comparison
# Built by hand from each fit's own stored fields, rather than via
# summarytable() (lcmm) -- summarytable() names its rows from the deparsed
# argument it's given, which would come out as the literal indexing
# expression ("stage1_fits[[\"K1\"]]") rather than a clean "K1" label when
# fits are pulled from a list like this, breaking the join below. BIC is
# computed directly (-2*loglik + npm*log(N)) rather than assumed to be a
# stored field, so this doesn't depend on exactly which summary fields a
# given lcmm version attaches to the fit object -- cross-check against
# summary(fit) once you have real fits, and use its BIC instead if it
# disagrees with the formula here.
diagnostics_row <- function(fit, K) {
  npm <- length(fit$best)
  N <- fit$ns  # number of subjects contributing to the fit
  bic <- -2 * fit$loglik + npm * log(N)

  class_diag <- if (K == 1) {
    tibble(entropy = NA_real_, mean_postprob = NA_real_)
  } else {
    pp <- postprob(fit)
    # pp$pprob's first two columns are id and modal class; the remaining
    # columns are per-class posterior probabilities. Confirm this layout
    # against str(pp$pprob) -- it hasn't been checked against a real fit.
    probs <- as.matrix(pp$pprob[, -(1:2), drop = FALSE])
    entropy <- 1 - mean(rowSums(-probs * log(probs + 1e-12)) / log(K))
    tibble(entropy = entropy, mean_postprob = mean(apply(probs, 1, max)))
  }

  tibble(K = K, loglik = fit$loglik, npm = npm, N = N,
    conv = fit$conv, BIC = bic) %>%
    bind_cols(class_diag)
}

diagnostics_summary <- purrr::imap_dfr(stage1_fits, function(fit, nm) {
  bind_cols(model_id = nm, diagnostics_row(fit, as.integer(sub("K", "", nm))))
})
```

```{r}
#| label: tbl-diagnostics-summary
kable(
  diagnostics_summary,
  caption = paste(
    "Convergence and classification-quality diagnostics across K.",
    "Lower BIC is better; entropy and mean posterior probability closer",
    "to 1 indicate cleaner class separation."
  )
)
```

```{r}
#| label: converged-models
# multlcmm's own convergence code is 1 for a proper local-optimum
# convergence; gridsearch() already selects the best-log-likelihood run
# across many starts for K >= 2, so a non-1 code surviving that process is
# a stronger signal of genuine difficulty than a single non-converged run
# would be.
converged_ids <- purrr::keep(stage1_fits, ~ .x$conv == 1) %>% names()
diagnostics_summary_converged <- diagnostics_summary %>% filter(model_id %in% converged_ids)
```

# Model comparison {#sec-comparison}

```{r}
#| label: bic-comparison
comparison_metrics <- diagnostics_summary_converged %>%
  arrange(BIC)
```

```{r}
#| label: tbl-comparison-overall
#| tbl-cap: "BIC and classification-quality diagnostics for all converged candidate models, sorted by BIC (lower is better)."
comparison_metrics %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kableExtra::kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

BIC alone can favor a larger $K$ that improves fit without actually
separating participants into distinct classes (@sec-methods). Before
accepting the BIC-minimizing $K$, check that its entropy and mean posterior
class-assignment probability (@tbl-diagnostics-summary) are also
reasonable -- there's no universal cutoff, but entropy well below, say,
0.6, or a mean assignment probability not much above $1/K$, is a sign that
$K$'s BIC improvement isn't coming from genuinely separated classes and is
worth weighing against a smaller, more cleanly separated $K$.

## Selecting the best-fitting model

```{r}
#| label: best-model
best_row <- comparison_metrics %>%
  slice_min(BIC, n = 1)

# model_id here is stage1_fits' own naming ("K1".."K4"; see
# stage1-comparison), not spec_grid's separate "multlcmm_K*" cache-file
# naming -- the two are different labels for the same K and shouldn't be
# confused when adding columns to either table later.
K_final <- as.integer(sub("^K", "", best_row$model_id))
fit_final <- stage1_fits[[paste0("K", K_final)]]

best_row
```

# Summary of the best-fitting model {#sec-summary}

```{r}
#| label: prep-final-fit
summary(fit_final)
```

## Parameter summary

```{r}
#| label: tbl-posterior-summary
#| tbl-cap: "Fixed-effect and variance-component estimates for the selected model (Wald SE/CI from the converged Hessian, not posterior draws)."
# summarytable()/summary() print to the console rather than returning a
# tidy object in every lcmm version; if summary(fit_final) above doesn't
# hand back something kable() can take directly, pull the coefficient
# table from fit_final$best and fit_final$V (the parameter vector and its
# variance-covariance matrix) instead -- both are always present on a
# converged fit regardless of summary()'s exact return type.
param_table <- tryCatch(
  as.data.frame(summary(fit_final)),
  error = function(e) {
    est <- fit_final$best
    se <- sqrt(diag(fit_final$V))
    tibble(parameter = names(est), estimate = est, se = se,
      wald_p = 2 * pnorm(-abs(est / se)))
  }
)

kable(param_table, digits = 3, booktabs = TRUE, longtable = TRUE)
```

## Class assignment (K \(\geq\) 2 models only)

```{r}
#| label: extract-class-assignments
# postprob() gives each participant's posterior class-membership
# probabilities and modal (hard) class assignment -- the ML-mixture
# analogue of the abandoned Stan model's per-draw class_prob, but a single
# point estimate per participant rather than a posterior distribution over
# assignments (there's no posterior here to characterize assignment
# uncertainty beyond the probability itself).
class_lookup <- if (K_final > 1) {
  pp <- postprob(fit_final)
  # pp$pprob's exact column names vary slightly by lcmm version; inspect
  # names(pp$pprob) if the rename() below errors. It should contain the
  # subject id (as passed via subject = "id"), the modal class number, and
  # one posterior-probability column per class.
  as_tibble(pp$pprob) %>%
    rename(id = 1, latent_class_num = class) %>%
    mutate(latent_class = factor(latent_class_num, levels = seq_len(K_final),
      labels = paste("Class", seq_len(K_final))))
} else {
  tibble(id = sort(unique(wide_qz$id)), latent_class = factor("Class 1"))
}

# Saved separately from fit_final itself (a small, portable id -> class
# lookup) specifically so a later session can subset ddl/ddl_qz by modal
# class and refit ltjm_general.stan (the K = 1 latent-time model, which
# converges without difficulty) within each class on its own -- the
# natural next step mentioned in the Introduction if this model converges
# cleanly. That refit is not implemented in this document.
dir.create("fits", showWarnings = FALSE)
saveRDS(class_lookup, file.path("fits", "class_lookup_final.RDS"))
```

## Baseline characteristics by class

```{r}
#| label: tbl-baseline-by-class
#| tbl-cap: "Characteristics of CLARiTI participants by modal latent class (best-fitting model)."
dd_subject <- ddl %>%
  select(id, NACCID, Dx, age0, RACE, SEX, EDUC, HISPANIC, APOE) %>%
  distinct()

dd_subject_class <- dd_subject %>%
  left_join(class_lookup %>% select(id, latent_class), by = "id", relationship = "one-to-one") %>%
  left_join(
    ddl %>%
      filter(!is.na(Y), !is.na(years)) %>%
      group_by(id, NACCID, Outcome) %>%
      slice_min(order_by = years, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(id, NACCID, Outcome, Y) %>%
      pivot_wider(names_from = Outcome, values_from = Y),
    by = c("id", "NACCID"), relationship = "one-to-one"
  )

gtsummary::tbl_summary(
  data = dd_subject_class,
  by = latent_class,
  include = c(Dx, age0, SEX, EDUC, RACE, HISPANIC, APOE,
    `Amyloid PET (CL)`, `Tau PET (SUVR)`, `Hipp. atrophy (Z-score)`, Memory, `Exec Function`),
  type = all_continuous() ~ "continuous2",
  statistic = list(
    all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
    all_categorical() ~ "{n} ({p}%)"
  ),
  digits = all_continuous() ~ 1,
  missing_text = "(Missing)"
) %>%
  add_overall(last = TRUE) %>%
  bold_labels()
```

## Population-level severity trajectories

```{r}
#| label: severity-trajectory-prep
# predictY() over an age grid, at reference covariate levels, is the
# multlcmm-native way to get fitted trajectories -- the frequentist
# analogue of the abandoned Stan model's posterior-draw trajectory
# summary (severity-quantile-prep), but from Monte Carlo draws around the
# asymptotic (Wald) parameter estimate rather than MCMC draws. Because
# @sec-methods-model's fixed/mixture terms are fully contrast()-ed (every
# outcome gets its own class-specific spline, not one shared latent
# trajectory), double-check predictY's returned column structure with
# names(pred_grid$pred) before trusting the reshape below -- this is a
# less common configuration than multlcmm's own single-latent-process
# vignette examples, and the exact column-naming convention for
# per-outcome x per-class predictions hasn't been confirmed by an actual
# run (the standing constraint on this project is that Claude never fits
# these models itself). If predictY doesn't split cleanly by outcome and
# class the way the code below assumes, fit_final$pred -- the fitted
# values at the observed data's own covariate patterns, always present on
# a converged fit regardless of how predictY handles this configuration --
# is the fallback: bin it by age and plot observed-vs-fitted directly
# instead of over a synthetic grid.
age_grid <- seq(min(wide_qz$age, na.rm = TRUE), max(wide_qz$age, na.rm = TRUE), length.out = 100)
ref_sex  <- names(sort(table(wide_qz$SEX), decreasing = TRUE))[1]
ref_educ <- mean(wide_qz$EDUC, na.rm = TRUE)

# The fitted model's formulas are in age_c (centered), not raw age
# (helper-multlcmm-fit), so newdata must supply age_c the same way,
# reusing the same age_center computed from wide_qz -- raw age is added
# back afterward purely for plotting.
newdata_grid <- tibble(age_c = age_grid - age_center, SEX = ref_sex, EDUC = ref_educ)

pred_grid <- predictY(fit_final, newdata = newdata_grid, var.time = "age_c", draws = TRUE)

# Reshape assuming column names of the form "<outcome>_class<k>" or
# similar; inspect names(pred_grid$pred) and adjust the regex below if the
# actual naming differs.
severity_summary_time <- as_tibble(pred_grid$pred, rownames = NULL) %>%
  mutate(age = age_grid) %>%
  tidyr::pivot_longer(-age, names_to = "col", values_to = "value") %>%
  tidyr::extract(col, into = c("outcome_lab", "stat", "k"),
    regex = "(Z[0-9]+).*class([0-9]+)", remove = FALSE) %>%
  mutate(outcome = as.integer(sub("Z", "", outcome_lab)), k = as.integer(k)) %>%
  left_join(outcome_map, by = "outcome") %>%
  mutate(
    latent_class = factor(k, levels = seq_len(K_final), labels = paste("Class", seq_len(K_final))),
    Outcome = recode(Outcome,
      "Amyloid PET (CL)" = "Amyloid PET", "Tau PET (SUVR)" = "Tau PET",
      "Hipp. atrophy (Z-score)" = "Hippocampal atrophy", "Exec Function" = "Executive function"),
    Outcome = factor(Outcome, levels = c("Amyloid PET", "Tau PET", "Hippocampal atrophy",
      "Memory", "Executive function"))
  )
```

```{r}
#| label: fig-severity-trajectory
#| fig-cap: "Modeled population-level severity trajectories by latent class over age, fixed covariates at reference values (best-fitting model)."
ggplot(severity_summary_time, aes(x = age, y = value, colour = latent_class, fill = latent_class)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ Outcome, ncol = 3) +
  scale_colour_brewer(palette = "Set1", name = "Latent class") +
  scale_fill_brewer(palette = "Set1", name = "Latent class") +
  labs(
    title = "Modeled population-level severity trajectories",
    subtitle = "Fitted mean trajectory by class; covariates fixed at reference values",
    x = "Age", y = "Predicted severity (Z-scale)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Observed participant-level trajectories

```{r}
#| label: fig-spaghetti-plot
#| fig-cap: "Observed participant-level severity trajectories by latent class (best-fitting model)."
spaghetti_data <- ddl_qz %>%
  left_join(class_lookup %>% select(id, latent_class), by = "id", relationship = "many-to-one") %>%
  mutate(
    Outcome = recode(Outcome,
      "Amyloid PET (CL)" = "Amyloid PET", "Tau PET (SUVR)" = "Tau PET",
      "Hipp. atrophy (Z-score)" = "Hippocampal atrophy", "Exec Function" = "Executive function"),
    Outcome = factor(Outcome, levels = c("Amyloid PET", "Tau PET", "Hippocampal atrophy",
      "Memory", "Executive function"))
  )

ggplot(spaghetti_data, aes(x = age, y = Q, group = interaction(id, Outcome), colour = latent_class)) +
  geom_line(alpha = 0.40, linewidth = 0.45) +
  geom_point(alpha = 0.35, size = 0.6) +
  facet_wrap(~ Outcome, ncol = 3) +
  scale_colour_brewer(palette = "Set1", name = "Latent class") +
  scale_y_continuous(breaks = seq(0, 1, 0.25)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Observed participant-level severity trajectories",
    subtitle = "Observed severity quantile plotted against chronological age",
    x = "Age", y = "Observed severity quantile"
  ) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Model fit checks

There is no posterior here to run classic Bayesian posterior-predictive
checks (`ppc_dens_overlay`, `ppc_stat`) against, the way the abandoned Stan
model's report did. The frequentist analogue available directly from a
converged `multlcmm` fit is a residual/fitted-value check: `fit_final$pred`
holds the fitted values at every observed data point (no synthetic grid
needed), which can be compared to the observed $Z$ values directly (e.g.,
binned by age, per outcome and class) to check for systematic
under/over-prediction that the class-specific spline trajectories above
might be masking at the population level.

# Limitations {#sec-limitations}

## The abandoned Bayesian latent-time joint model (Stan)

Before adopting the `multlcmm` approach used throughout this document, an
extended effort built two general-purpose Stan programs
(`ltjm_general.stan` for $K=1$, `lctjm_general.stan` for $K \ge 2$) around
a participant-specific latent time shift $\delta_i$, fit via CmdStanR (4
chains), and swept a grid over disease-time scale (chronological age vs.
time since first observation), fixed-effect intercept structure
(outcome-specific vs. common), class-specific intercept structure, and
outcome weighting. That approach converged cleanly at $K \le 2$ under
every configuration tried, but never reached acceptable convergence at
$K = 3$, across a long sequence of identification schemes for the
class-specific intercepts $\alpha_{jk}$, each motivated by the specific
failure of the one before it:

1. A single hard-zero intercept constraint on the same outcome
   (hippocampal atrophy) already ordering the class-defining slope. This
   converged at low $K$ but forced that outcome's fitted class means to
   coincide exactly at disease-time zero, producing visibly artificial
   trajectory intersections.
2. A sum-to-zero constraint across outcomes within each class
   ($\sum_j \alpha_{jk} = 0$), which avoided the artificial intersection
   but left every outcome's intercept completely unordered across classes.
   $K = 3$ failed to converge, with weak separation in the one ordered
   dimension (the slope anchor) letting the unordered intercepts swap
   between classes without technically violating any constraint.
3. Full ordering of every outcome's intercept (and, briefly, every
   outcome's slope too) across classes. This measurably regressed $K = 2$
   (previously clean; max_rhat 1.734, 130 parameters above $\hat R=1.01$
   under full slope ordering) and left $K=3$/$K=4$ comparably broken --
   zero divergent transitions alongside enormous Rhat/ESS, the textbook
   signature of chains trapped in different local optima
   (`diagnose_mode_trapping.R`) rather than a boundary or geometry problem.
4. A two-anchor hard-zero scheme: hippocampal atrophy's slope orders the
   classes as before, but a *different* outcome (memory) gets the
   hard-zero intercept constraint, avoiding the single-outcome overload of
   scheme 1. Combined with recentering the age axis at the 5th percentile
   of baseline age (rather than the sample mean) to relocate the resulting
   forced-trajectory-coincidence point from an arbitrary mid-course age to
   cohort entry -- a defensible assumption in growth-mixture modeling, not
   just a cosmetic fix, since $\delta_i$'s prior is centered at 0
   regardless of where the age axis itself is centered. This was the most
   scientifically defensible scheme tried and still did not converge at
   $K=3$: max_rhat 2.947, max_feff_rhat 2.947, 4285 parameters above
   $\hat R = 1.01$, 0 divergences.
5. Removing class-specific intercepts entirely
   ($\alpha_{jk} \equiv 0$ for every outcome and class), leaving only
   hippocampal atrophy's slope ordering to distinguish classes. This
   removes the whole identifiability sub-problem rather than proposing a
   sixth parameterization of it, and was the last step taken before this
   document moved to `multlcmm`; whether it resolves $K=3$ convergence on
   its own was not established before the switch.

Every step changed either the fitted convergence or the substantive claim
being made, which is why the mechanism, not just the outcome, is preserved
here rather than only the final negative result. `lctjm_general.stan`
still implements every scheme in the sequence (governed by data flags), so
any of them can be revisited without rewriting the Stan program, should a
reason arise to return to a latent-time joint-model framework. The
diagnostic scripts built to investigate this (`diagnose_mode_trapping.R`,
`explore_k3_modes.R`) remain useful for characterizing multi-modality in
any future MCMC-based model on this data.

## Limitations of the current (`multlcmm`) approach

- **Local optima.** `gridsearch()`'s 50 random starts per $K \ge 2$ reduce
  but don't eliminate the risk of settling on a local rather than global
  optimum (@sec-methods-optima); a converged, BIC-competitive $K$ that
  looks qualitatively different across separate `gridsearch()` runs with a
  different seed or start count is worth another look before treating it
  as final.
- **The single-shared-latent-process assumption is only partly relaxed.**
  Wrapping every outcome's spline term in `contrast()` gives each outcome
  its own class-specific trajectory shape (@sec-methods-model), but the
  two subject-level random effects tying the $J$ outcomes together for a
  given participant ($u_i$, $v_i$) are still shared, not outcome-specific,
  beyond the outcome-specific random intercept $w_{ij}$. A participant who
  runs persistently faster or slower than their class average does so by
  the same amount, in the same direction, on every outcome's spline
  coefficients (modulated only by each outcome's own loading), which is a
  narrower within-subject correlation structure than a fully general
  multivariate mixed model would allow. Confirm this against
  `summary(fit_final)`/`fit_final$Xnames` once fit, since it wasn't
  possible to verify by execution here.
- **BIC/entropy is a different paradigm from WAIC/LOOIC, not a drop-in
  replacement.** Model-selection numbers reported here (@sec-comparison)
  are not comparable in magnitude or interpretation to the WAIC/LOOIC
  values the abandoned Stan grid would have produced; don't carry over any
  intuition about "how much better" one $K$ is from that earlier framework.
- **Spline degrees of freedom fixed, not swept.** 3 df was chosen directly
  rather than compared against 2 df or other choices (@sec-grid); if the
  fitted class-specific trajectories look overfit or underfit for a given
  outcome, revisiting this as a genuine comparison axis is a reasonable
  next step.
- **No time-to-event component.** This is by design (@sec-methods), not an
  oversight -- `multlcmm` does not support one, and this analysis has no
  survival/censoring outcome to link to one anyway.
- **The latent-time refit per class (if this converges) is not yet
  implemented.** @sec-summary saves a small id-to-class lookup
  specifically to support subsetting `ddl`/`ddl_qz` by modal class and
  refitting `ltjm_general.stan` within each one afterward, but that refit
  itself is future work, not part of this document.

# Session information

```{r}
#| label: session-info
sessionInfo()
```
